### 2022-04-25 main.tex
Initial Overleaf Import

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Adversarial Penguin}
\author{Steve K Harris}
\date{April 2022}

\begin{document}

\maketitle

\section{Introduction}

\end{document}

### 2022-04-25 main.tex
accepted abstract loaded; switched to markdown

@@
-\documentclass{article}
-\usepackage[utf8]{inputenc}
+% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{Adversarial Penguin}
-\author{Steve K Harris}
-\date{April 2022}
+%%% Version 3.4 Generated 2018/06/15 %%%
+%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
+
+\documentclass[utf8]{FrontiersinVancouver}
+
+\usepackage{url,hyperref,lineno,microtype,subcaption}
+\usepackage[onehalfspacing]{setspace}
+\usepackage[footnotes,definitionLists,hashEnumerators,smartEllipses,hybrid,citations]{markdown}
+
+\linenumbers
+
+% Leave a blank line between paragraphs instead of using \\
+
+\def\keyFont{\fontsize{8}{11}\helveticabold }
+\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
+\def\Authors{First Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^{1,2}$}
+% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
+% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
+\def\Address{$^{1}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X \\
+$^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
+% The Corresponding Author should be marked with an asterisk
+% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
+\def\corrAuthor{Corresponding Author}
+
+\def\corrEmail{email@uni.edu}
 
 \begin{document}
+\onecolumn
+\firstpage{1}
+
+\title[Running Title]{Article Title}
+
+\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
+\address{} %This field will be automatically populated
+\correspondance{} %This field will be automatically populated
+
+\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
+%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}
 
@@
 
+\begin{abstract}
+
+%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
+\section{}
+\markdownInput{00-abstract.md}
+
+For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.
+
+As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership.
+References should not be cited in the abstract.
+Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.
+
+\tiny
+ \keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+\end{abstract}
+
 \section{Introduction}
+\markdownInput{01-main.md}
+
+For Original Research Articles \citep{conference}, Clinical Trial Articles \citep{article}, and Technology Reports \citep{patent}, the introduction should be succinct, with no subheadings \citep{book}.
+For Case Reports the Introduction should include symptoms at presentation \citep{chapter}, physical exams and lab results \citep{dataset}.
+
+\section{Article types}
+
+For requirements for a specific article type please refer to the Article Types on any Frontiers journal page.
+Please also refer to  \href{http://home.frontiersin.org/about/author-guidelines#Sections}{Author Guidelines} for further information on how to organize your manuscript in the required sections or their equivalents for your field
+
+% For Original Research articles, please note that the Material and Methods section can be placed in any of the following ways: before Results, before Discussion or after Discussion.
+
+\section{Manuscript Formatting}
+
+\subsection{Heading Levels}
+
+%There are 5 heading levels
+
+\subsection{Level 2}
+\subsubsection{Level 3}
+\paragraph{Level 4}
+\subparagraph{Level 5}
+
+\subsection{Equations}
+Equations should be inserted in editable format from the equation editor.
+
+\begin{equation}
+\sum x+ y =Z\label{eq:01}
+\end{equation}
+
+\subsection{Figures}
+Frontiers requires figures to be submitted individually, in the same order as they are referred to in the manuscript.
+Figures will then be automatically embedded at the bottom of the submitted manuscript.
+Kindly ensure that each table and figure is mentioned in the text and in numerical order.
+Figures must be of sufficient resolution for publication \href{https://www.frontiersin.org/about/author-guidelines#ImageSizeRequirements}{see here for examples and minimum requirements}.
+Figures which are not according to the guidelines will cause substantial delay during the production process.
+Please see \href{https://www.frontiersin.org/about/author-guidelines#FigureRequirementsStyleGuidelines}{here} for full figure guidelines.
+Cite figures with subfigures as figure \ref{fig:2}B.
+
+\subsubsection{Permission to Reuse and Copyright}
+Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet).
+It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.
+%%Figures, tables, and images will be published under a Creative Commons CC-BY licence and permission must be obtained for use of copyrighted material from other sources (including re-published/adapted/modified/partial figures and images from the internet).
+It is the responsibility of the authors to acquire the licenses, to follow any citation instructions requested by third-party rights holders, and cover any supplementary charges.
+
+\subsection{Tables}
+Tables should be inserted at the end of the manuscript.
+Please build your table directly in LaTeX.Tables provided as jpeg/tiff files will not be accepted.
+Please note that very large tables (covering several pages) cannot be included in the final PDF for reasons of space.
+These tables will be published as \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} on the online article page at the time of acceptance.
+The author will be notified during the typesetting of the final article if this is the case.
+
+\section{Nomenclature}
+
+\subsection{Resource Identification Initiative}
+To take part in the Resource Identification Initiative, please use the corresponding catalog number and RRID in your current manuscript.
+For more information about the project and for steps on how to search for an RRID, please click \href{http://www.frontiersin.org/files/pdf/letter_to_author.pdf}{here}.
+
+\subsection{Life Science Identifiers}
+Life Science Identifiers (LSIDs) for ZOOBANK registered names or nomenclatural acts should be listed in the manuscript before the keywords.
+For more information on LSIDs please see \href{https://www.frontiersin.org/about/author-guidelines#Nomenclature}{Inclusion of Zoological Nomenclature} section of the guidelines.
+
+\section{Additional Requirements}
+
+For additional requirements for specific article types and further information please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines#AdditionalRequirements}{Author Guidelines}.
+
+\section*{Conflict of Interest Statement}
+%All financial, commercial or other relationships that might be perceived by the academic community as representing a potential conflict of interest must be disclosed.
+If no such relationship exists, authors will be asked to confirm the following statement:
+
+The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
+
+\section*{Author Contributions}
+
+The Author Contributions section is mandatory for all articles, including articles by sole authors.
+If an appropriate statement is not provided on submission, a standard one will be inserted during the production process.
+The Author Contributions statement must describe the contributions of individual authors referred to by their initials and, in doing so, all authors agree to be accountable for the content of the work.
+Please see  \href{https://www.frontiersin.org/about/policies-and-publication-ethics#AuthorshipAuthorResponsibilities}{here} for full authorship criteria.
+
+\section*{Funding}
+Details of all funding sources should be provided, including grant numbers if applicable. Please ensure to add all necessary funding information, as after publication this is no longer possible.
+
+\section*{Acknowledgments}
+This is a short text to acknowledge the contributions of specific colleagues, institutions, or agencies that aided the efforts of the authors.
+
+\section*{Supplemental Data}
+ \href{http://home.frontiersin.org/about/author-guidelines#SupplementaryMaterial}{Supplementary Material} should be uploaded separately on submission, if there are Supplementary Figures, please include the caption in the same file as the figure.
+LaTeX Supplementary Material templates can be found in the Frontiers LaTeX folder.
+
+\section*{Data Availability Statement}
+The datasets [GENERATED/ANALYZED] for this study can be found in the [NAME OF REPOSITORY] [LINK].
+% Please see the availability of data guidelines for more information, at https://www.frontiersin.org/about/author-guidelines#AvailabilityofData
+
+\bibliographystyle{Frontiers-Vancouver} 
+\bibliography{test}
+
+%%% Make sure to upload the bib file along with the tex file and PDF
+%%% Please see the test.bib file for some examples of references
+
+\section*{Figure captions}
+
+%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
+%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
+%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
+%%% If using *.tif files convert them to .jpg or .png
+%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=10cm]{logo1}% This is a *.eps file
+\end{center}
+\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=15cm]{logos}
+\end{center}
+\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
+\end{figure}
+
+%%% If you are submitting a figure with subfigures please combine these into one image file with part labels integrated.
+%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
+%%% Frontiers will add the figures at the end of the provisional pdf automatically
+%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.
 

### 2022-04-29 main.tex
authors etc

--- 2022-04-25 main.tex
@@
-\documentclass{article}
-\usepackage[utf8]{inputenc}
+% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{Adversarial Penguin}
-\author{Steve K Harris}
-\date{April 2022}
+%%% Version 3.4 Generated 2018/06/15 %%%
+%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
+
+\documentclass[utf8]{FrontiersinVancouver}
+
+\usepackage{url,hyperref,lineno,microtype,subcaption}
+\usepackage[onehalfspacing]{setspace}
+\usepackage[footnotes,definitionLists,hashEnumerators,smartEllipses,hybrid,citations]{markdown}
+
+\linenumbers
+
+% Leave a blank line between paragraphs instead of using \\
+
+\def\keyFont{\fontsize{8}{11}\helveticabold }
+\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
+\def\Authors{Steve Harris\,$^{1,3*}$, Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Tim Bonnici \,$^{1,3}$ and Nel Swanepoel\,$^{2}$}
+% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
+% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
+\def\Address{
+$^{1}$Institute of Health Informatics, University College London, UK \\
+$^{2}$Centre for Advanced Research Computing, University College London, UK \\
+$^{3}$Department of Critical Care, University College London Hospital, UK \\
+}
+% $^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
+% The Corresponding Author should be marked with an asterisk
+% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
+\def\corrAuthor{Corresponding Author}
+
+\def\corrEmail{steve.harris@ucl.ac.uk}
 
 \begin{document}
+\onecolumn
+\firstpage{1}
+
+\title[Running Title]{Article Title}
+
+\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
+\address{} %This field will be automatically populated
+\correspondance{} %This field will be automatically populated
+
+\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
+%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}
 
@@
 
-\section{Introduction}
+\begin{abstract}
+
+%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
+\section{}
+\markdownInput{00-abstract.md}
+
+For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.
+
+As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership.
+References should not be cited in the abstract.
+Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.
+
+\tiny
+ \keyFont{ \section{Keywords:} translational medicine, machine learning, health informatics, ml-ops, safety, artificial intelligence} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+\end{abstract}
+
+\markdownInput{01-main.md}
+
+\section*{Conflict of Interest Statement}
+%All financial, commercial or other relationships that might be perceived by the academic community as representing a potential conflict of interest must be disclosed.
+If no such relationship exists, authors will be asked to confirm the following statement:
+
+The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
+
+\section*{Author Contributions}
+All authors listed have made a substantial, direct and intellectual contribution to the work, and approved it for publication.
+
+\section*{Funding}
+
+SH is supported by a Health Foundation Improvement Science Fellowship, and by funding from 
+Biomedical Research Centre based at the University College Hospitals (UCLH) National Health Service (NHS) Foundation Trust and University College London (UCL).
+SH, TK, and NS are supported funds from the National Institute for Health Research (Artificial Intelligence, Digitally adapted, hyper-local realtime bed forecasting to manage flow for NHS wards, AI_AWARD01786) and NHS-X.
+The views expressed in this publication are those of the author(s) and not necessarily those of the National Institute for Health Research, NHS-X or the Department of Health and Social Care.
+WL is supported by a
+
+\bibliographystyle{Frontiers-Vancouver} 
+\bibliography{test}
+
+%%% Make sure to upload the bib file along with the tex file and PDF
+%%% Please see the test.bib file for some examples of references
+
+\section*{Figure captions}
+
+%%% Please be aware that for original research articles we only permit a combined number of 15 figures and tables, one figure with multiple subfigures will count as only one figure.
+%%% Use this if adding the figures directly in the mansucript, if so, please remember to also upload the files when submitting your article
+%%% There is no need for adding the file termination, as long as you indicate where the file is saved. In the examples below the files (logo1.eps and logos.eps) are in the Frontiers LaTeX folder
+%%% If using *.tif files convert them to .jpg or .png
+%%%  NB logo1.eps is required in the path in order to correctly compile front page header %%%
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=10cm]{figs/emap.png}% This is a *.eps file
+\end{center}
+\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics[width=15cm]{figs/flowehr.png}
+\end{center}
+\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
+\end{figure}
+
+%%% If you are submitting a figure with subfigures please combine these into one image file with part labels integrated.
+%%% If you don't add the figures in the LaTeX files, please upload them when submitting the article.
+%%% Frontiers will add the figures at the end of the provisional pdf automatically
+%%% The use of LaTeX coding to draw Diagrams/Figures/Structures should be avoided. They should be external callouts including graphics.
 

### 2022-04-29 template-snippets.tex
authors etc

--- 2022-04-25 main.tex
@@
 \def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{First Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^{1,2}$}
+\def\Authors{Steve Harris\,$^{1,3*}$, Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Tim Bonnici \,$^{1,3}$ and Nel Swanepoel\,$^{2}$}
 % Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
 % If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
-\def\Address{$^{1}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X \\
-$^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
+\def\Address{
+$^{1}$Institute of Health Informatics, University College London, UK \\
+$^{2}$Centre for Advanced Research Computing, University College London, UK \\
+$^{3}$Department of Critical Care, University College London Hospital, UK \\
+}
+% $^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
 % The Corresponding Author should be marked with an asterisk
@@
 
-\def\corrEmail{email@uni.edu}
+\def\corrEmail{steve.harris@ucl.ac.uk}
 
@@
 \tiny
- \keyFont{ \section{Keywords:} keyword, keyword, keyword, keyword, keyword, keyword, keyword, keyword} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
+ \keyFont{ \section{Keywords:} translational medicine, machine learning, health informatics, ml-ops, safety, artificial intelligence} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
 \end{abstract}

### 2022-04-30 main.tex
bibliography update

@@
 \def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{Steve Harris\,$^{1,3*}$, Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Tim Bonnici \,$^{1,3}$ and Nel Swanepoel\,$^{2}$}
+\def\Authors{Steve Harris\,$^{1,*}$, Thomas Keen\,$^{2}$ and Nel Swanepoel\,$^{1,2}$}
 % Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
 % If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
-\def\Address{
-$^{1}$Institute of Health Informatics, University College London, UK \\
-$^{2}$Centre for Advanced Research Computing, University College London, UK \\
-$^{3}$Department of Critical Care, University College London Hospital, UK \\
-}
-% $^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
+\def\Address{$^{1}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X \\
+$^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
 % The Corresponding Author should be marked with an asterisk
@@
 
-\def\corrEmail{steve.harris@ucl.ac.uk}
+\def\corrEmail{email@uni.edu}
 
@@
 
-\title[Running Title]{Article Title}
+\title[Translational Digital Medicine]{The Five Pillars of Translational Digital Medicine}
 
@@
 \maketitle
-
 \begin{abstract}
@@
 
-For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.
+\tiny
+ \keyFont{ \section{Keywords:} translational medicine, machine learning, health informatics, ml-ops, safety, artificial intelligence} 
+ %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
 
-As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership.
-References should not be cited in the abstract.
-Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.
-
-\tiny
- \keyFont{ \section{Keywords:} translational medicine, machine learning, health informatics, ml-ops, safety, artificial intelligence} %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
 \end{abstract}
@@
 Biomedical Research Centre based at the University College Hospitals (UCLH) National Health Service (NHS) Foundation Trust and University College London (UCL).
-SH, TK, and NS are supported funds from the National Institute for Health Research (Artificial Intelligence, Digitally adapted, hyper-local realtime bed forecasting to manage flow for NHS wards, AI_AWARD01786) and NHS-X.
+SH, TK, and NS are supported funds from the National Institute for Health Research (Artificial Intelligence, Digitally adapted, hyper-local real-time bed forecasting to manage flow for NHS wards, AI AWARD01786) and NHS-X.
 The views expressed in this publication are those of the author(s) and not necessarily those of the National Institute for Health Research, NHS-X or the Department of Health and Social Care.
@@
 \bibliographystyle{Frontiers-Vancouver} 
-\bibliography{test}
+\bibliography{refs}
 

### 2022-04-30 main.tex
removed dataset from misc bibitem

@@
 \def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{Steve Harris\,$^{1,*}$, Thomas Keen\,$^{2}$ and Nel Swanepoel\,$^{1,2}$}
+\def\Authors{Steve Harris\,$^{1,2*}$, Thomas Keen\,$^{1}$ Tim Bonnici\,$^{1,2}$ Watjana Lilaonitkul\,$^{1}$ and Nel Swanepoel\,$^{3}$}
 % Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
 % If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
-\def\Address{$^{1}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X \\
-$^{2}$Laboratory X, Institute X, Department X, Organization X, City X , State XX (only USA, Canada and Australia), Country X  }
+
+\def\Address{$^{1}$Institute of Health Informatics, University College London, London, UK \\
+$^{2}$Department of Critical Care, University College London Hospital, London, UK  \\
+$^{3}$Centre for Advanced Research Computing, University College London, London, UK  }
+
 % The Corresponding Author should be marked with an asterisk

### 2022-04-30 main.tex
discussion and edits from nel

@@
 \def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{Steve Harris\,$^{1,2*}$, Thomas Keen\,$^{1}$ Tim Bonnici\,$^{1,2}$ Watjana Lilaonitkul\,$^{1}$ and Nel Swanepoel\,$^{3}$}
+\def\Authors{Steve Harris\,$^{1,2*}$, Tim Bonnici\,$^{1,2}$ Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Nel Swanepoel\,$^{3}$ Mark White\,$^{4}$}
 % Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
@@
 $^{2}$Department of Critical Care, University College London Hospital, London, UK  \\
-$^{3}$Centre for Advanced Research Computing, University College London, London, UK  }
+$^{3}$Centre for Advanced Research Computing, University College London, London, UK \\
+$^{4}$Digital Healthcare, University College London Hospital, London, UK   }
 
@@
 Biomedical Research Centre based at the University College Hospitals (UCLH) National Health Service (NHS) Foundation Trust and University College London (UCL).
+NS is supported by funding from Health Data Research UK.
 SH, TK, and NS are supported funds from the National Institute for Health Research (Artificial Intelligence, Digitally adapted, hyper-local real-time bed forecasting to manage flow for NHS wards, AI AWARD01786) and NHS-X.
 The views expressed in this publication are those of the author(s) and not necessarily those of the National Institute for Health Research, NHS-X or the Department of Health and Social Care.
-WL is supported by a
+WL is supported by a UKRI Ernest Rutherford Fellowship.
 

### 2022-04-30 main.tex
abstract

@@
 
-SH is supported by a Health Foundation Improvement Science Fellowship, and by funding from 
-Biomedical Research Centre based at the University College Hospitals (UCLH) National Health Service (NHS) Foundation Trust and University College London (UCL).
+SH is supported by a Health Foundation Improvement Science Fellowship, and by funding from  the Biomedical Research Centre based at the University College Hospitals (UCLH) National Health Service (NHS) Foundation Trust and University College London (UCL).
 NS is supported by funding from Health Data Research UK.
 SH, TK, and NS are supported funds from the National Institute for Health Research (Artificial Intelligence, Digitally adapted, hyper-local real-time bed forecasting to manage flow for NHS wards, AI AWARD01786) and NHS-X.
+WL is supported by a UKRI Ernest Rutherford Fellowship.
 The views expressed in this publication are those of the author(s) and not necessarily those of the National Institute for Health Research, NHS-X or the Department of Health and Social Care.
-WL is supported by a UKRI Ernest Rutherford Fellowship.
 

### 2022-05-08 main.tex
waty's work

@@
 \def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{Steve Harris\,$^{1,2*}$, Tim Bonnici\,$^{1,2}$ Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Nel Swanepoel\,$^{3}$ Mark White\,$^{4}$}
+\def\Authors{Steve Harris\,$^{1,2*}$ Tim Bonnici\,$^{1,2}$ Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Mark J White\,$^{4}$ Nel Swanepoel\,$^{3}$ }
 % Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).

### 2022-05-08 main.tex
captions and ESM

@@
 
-\markdownInput{01-main.md}
+% \markdownInput{01-main.md}
+\markdownInput{01-intro.md}
+\markdownInput{02-real-world-development.md}
+\markdownInput{03-ml-ops.md}
+\markdownInput{04-responsible-ai.md}
+\markdownInput{05-implementation.md}
+\markdownInput{06-continuous-evaluation.md}
+\markdownInput{08-discussion.md}
 
@@
 \end{center}
-\caption{ Enter the caption for your figure here.  Repeat as  necessary for each of your figures}\label{fig:1}
+\caption{ 
+Our real-world development is performed on the  Experimental Medicine Application Platform (EMAP).
+EMAP is a clinical laboratory within which ML4H researchers can iteratively build, test and gather feedback from the bedside.
+It unifies the data and the tools for off-line and online development of ML4H models.
+In brief, EMAP builds a patient orientated SQL database from HL7 version 2 (HL7v2) messages that are being exchanged between hospital systems.
+HL7v2 messages are ubiquitous in health care, and the \textit{de facto} standard for internal communication.
+}\label{fig:1}
 \end{figure}
@@
 \end{center}
-\caption{This is a figure with sub figures, \textbf{(A)} is one logo, \textbf{(B)} is a different logo.}\label{fig:2}
+\caption{
+Our ML-Ops platform is called FlowEHR.
+Moving from left to right across the figure, the system monitors raw input data including checks for distribution shift, builds features with testable and quality controlled code, makes those features available to for both training and predictions to avoid train/serve skew, and maintains an auditable and monitored model repository.
+}\label{fig:2}
 \end{figure}
@@
 
+\section*{Electronic Supplementary Material}
+\markdownInput{95-supplementary-material.md}
+
 \end{document}

### 2022-05-08 main.tex
frontiers esm approach
working pre-amble

@@
 \def\keyFont{\fontsize{8}{11}\helveticabold }
-\def\firstAuthorLast{Sample {et~al.}} %use et al only if is more than 1 author
-\def\Authors{Steve Harris\,$^{1,2*}$ Tim Bonnici\,$^{1,2}$ Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Mark J White\,$^{4}$ Nel Swanepoel\,$^{3}$ }
+\def\firstAuthorLast{Harris {et~al.}} %use et al only if is more than 1 author
+\def\Authors{Steve Harris\,$^{1,2*}$\textsuperscript{\textdagger} Tim Bonnici\,$^{1,2}$\textsuperscript{\textdagger} Thomas Keen\,$^{1}$ Watjana Lilaonitkul\,$^{1}$ Mark J White\,$^{4}$ Nel Swanepoel\,$^{3}$ }
 % Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
@@
 \maketitle
+\textsuperscript{\textdagger}These authors have contributed equally to this work and share first authorship\\
+
 \begin{abstract}
@@
 % \markdownInput{01-main.md}
-\markdownInput{01-intro.md}
-\markdownInput{02-real-world-development.md}
-\markdownInput{03-ml-ops.md}
-\markdownInput{04-responsible-ai.md}
-\markdownInput{05-implementation.md}
-\markdownInput{06-continuous-evaluation.md}
-\markdownInput{08-discussion.md}
+% \markdownInput{01-intro.md}
+% \markdownInput{02-real-world-development.md}
+% \markdownInput{03-ml-ops.md}
+% \markdownInput{04-responsible-ai.md}
+% \markdownInput{05-implementation.md}
+% \markdownInput{06-continuous-evaluation.md}
+% \markdownInput{08-discussion.md}
 
@@
 \begin{center}
-\includegraphics[width=10cm]{figs/emap.png}% This is a *.eps file
+\includegraphics[width=10cm]{assets/emap.png}% This is a *.eps file
 \end{center}
@@
 \begin{center}
-\includegraphics[width=15cm]{figs/flowehr.png}
+\includegraphics[width=15cm]{assets/flowehr.png}
 \end{center}
@@
 Our ML-Ops platform is called FlowEHR.
-Moving from left to right across the figure, the system monitors raw input data including checks for distribution shift, builds features with testable and quality controlled code, makes those features available to for both training and predictions to avoid train/serve skew, and maintains an auditable and monitored model repository.
-}\label{fig:2}
+Moving from left to right across the figure, the system monitors raw input data including checks for distribution shift, builds features with testable and quality controlled code, makes those features available to for both training and predictions to avoid train/serve skew, and maintains an auditable and monitored model repository.}\label{fig:2}
 \end{figure}
@@
 
-\section*{Electronic Supplementary Material}
-\markdownInput{95-supplementary-material.md}
-
 \end{document}

### 2022-06-27 rebuttals/response-reviewer-3.md
reviewer 3 - major 1

*************************  
Independent Review Report, Reviewer 3  
EVALUATION  
Please list your revision requests for the authors and provide your detailed comments, including highlighting limitations and strengths of the review. If you have additional comments based on Q2 and Q3 you can add them as well.  
Summary: I congratulate the authors on writing a wonderful overview on how hospitals can reach the digital maturity to deploy machine learning algorithms to support clinical decision making. The authors describe 5 major pillars that form a Clinical Deployment Environment. The figures also provide readers an idea of the major (and many) components that are necessary to erect a real-world CDEs. Finally, the paper discusses phased trials for testing ML algorithms.

Major:

1. I am confused what the terminology “code-to-data” means exactly. The authors never really provide a definition of what it means to reverse the typical setup of “data-to-code”. My first impression was that it meant the ML team would write code that would generate data. However, this does not seem to be the message of the first pillar. Instead, the message seems to be that there should be infrastructure for accessing live clinical data without harming the live clinical system. Please remove/replace this terminology to avoid reader confusion.

2. The second pillar describes a team in the hospital for ML-Ops. This idea has been described in a number of papers, including Cosgrif 2020, Eaneff 2020, and Feng 2022. These papers have drawn parallels to existing infrastructure in the hospital to show that the idea of MLOps is not foreign and to bring urgency and relevance. For instance, Feng 2022 describes interesting parallels between hospital Quality Improvement units and ML-Ops team (they called this an AI-QI team). Eaneff 2020 refers to such teams as “algorithm stewards” to draw links to the concept of “hospital stewards” that is well-understood in the clinical space. I think the case for Pillar 2 would be stronger if the authors mention these connections.

3. Recent work have questioned the value of model explainability, particularly because many explainability methods exhibit flaws. See Ghassemi, Oakden-Rayner, and Beam 2021 and [https://arxiv.org/abs/2202.01602](https://arxiv.org/abs/2202.01602), as well as references within. I think the recommendation for model explainability should be qualified with this fact that the current explainability methods leave much to be desired.

4. There is a quickly growing literature on fail-safes in clinical ML and the authors should point readers to it. This idea has been referred to as selective prediction (Chow, 1970; Tortorella,2000; Bartlett and Wegkamp, 2008; El-Yaniv and Wiener, 2010; Feng, Sondhi, Perry, and Simon 2021). The idea of deploying ensembles with fail-safes is promising as well, and has also been previously described in Feng 2021.

5. While I agree that dynamic model calibration is an important tool to maintain the reliability of ML systems, the first motivating example provided in Section 3.3 seem appropriate. The example states that an ML model can influence shifts in its input data, which can lead to performance deterioration. However, the authors also cite Liley, which shows an example where there are shifts induced by the model but the model itself should not be updated. The second motivating example (models trained on a bad dataset and therefore should be updated) is a good reason for model updating, though dynamic model calibration may not be sufficient in this case; instead larger model revisions may be required. Perhaps a better motivating example for dynamic model calibration is provided in Davis, Lasko, Chen, Siew, and Matheny 2017: They show how a ML model for acute kidney injury decays in model calibration.

6. The authors do not provide references for how to perform dynamic model calibration. The following papers have proposed methods for doing so, and should be included in the manuscript: Davis, Greevy, Lasko, Walsh, Matheny 2020 and Feng, Gossmann, Sahiner, and Pirracchio 2022.

7. It is very interesting that the authors were able to implement continuous model evaluation using encouragement-based randomization within their hospital infrastructure. It would be useful for the authors to provide additional explanations on the ethical considerations and why randomization without consent is justified. The authors should also highlight the complexity of estimating the ML model’s impact in encouragement designs due to selection bias and noncompliance: see e.g. Barnard 2011 and Frangakis et al 2002.

8. The idea of phased trials to test prediction models in medicine is not new. Similar ideas have been previously described in Pepe et.al. 2001 and Pepe 2005. That said, there are still some differences in the definition of the phases described in the Pepe papers versus those described in the paper, because the former was focused on models that did not rely on EHR data. The authors should provide additional context by comparing their ideas to those previously described in these papers.

9. It is not clear to me how the phased trials for testing algorithms relate to the Clinical Deployment Environment. How is the CDE used in each of the phases (if at all)? Which of the pillars are relevant in each of the phases? For instance, there is debate on which phase of model development we should investigate issues of AI safety and fairness (Wawira et al 2021). Please connect this section with the rest of the manuscript. Finally, continuous model evaluation (Pillar 5) could also be viewed as a Phase IV trial, which is common in drug regulation for monitoring adverse events after drug approval is granted.

10. Figure 1 has a lot of components, and it is not clear what all of the components do. Please expand the caption to explain the major components in the figure. For instance, what is the “immutable data store” versus the “user data store”? What is the “message queue” vs the “message reader”? Also, what value add does the reporting DB offer over the live EHR?

Minor:  
1. Please remove the use of undefined acronyms in the figures (e.g. HL7 and HDR-UK) or provide definitions/citations.  
Check List  
a. Is the quality of the figures and tables satisfactory?  
Yes  
b. Does the reference list cover the relevant literature adequately and in an unbiased manner?  
Yes  
c. Does this manuscript refer only to published data? (unpublished or original data is not allowed for this article type)  
No  
d. Does the review include a balanced, comprehensive, and critical view of the research area?  
Yes  
QUALITY ASSESSMENT:  
Rigor  
3  
Quality of the writing  
5  
Overall quality of the content  
5  
Interest to a general audience  
5

### 2022-06-27 rebuttals/response-reviewer-3.md
reviewer 3 (except Waty's issues)

@@
-*************************  
-Independent Review Report, Reviewer 3  
-EVALUATION  
-Please list your revision requests for the authors and provide your detailed comments, including highlighting limitations and strengths of the review.
-If you have additional comments based on Q2 and Q3 you can add them as well.
-Summary: I congratulate the authors on writing a wonderful overview on how hospitals can reach the digital maturity to deploy machine learning algorithms to support clinical decision making.
+> Summary: I congratulate the authors on writing a wonderful overview on how hospitals can reach the digital maturity to deploy machine learning algorithms to support clinical decision making.
 The authors describe 5 major pillars that form a Clinical Deployment Environment.
@@
 
-Major:
+Thank you for these comments. This was exactly our aim: to provide an schematic overview of the steps that will be required to reach digital maturity given our own experience.
+
+## Major:
 
@@
 
+*We have edited the text to make this clearer.
+In brief, Guinney et al coined the phrase 'model to data' (ref below) to describe an inversion of the concept of data sharing to improve confidentiality.
+Here the model (code) is imported to the source data environment rather than source data being shared with external teams.
+Confidentialty and privacy improves because data is not copied and shared.
+We take this further and argue that 'data-to-code' ('data to modeler' in their parlance) is flawed because of the separation it imposes between the modellers and the end users.*
+
+*Concretely under Section 1 (Real World Development) we have replaced paragaphs 2-4 with*
+
+> Given the sensitive nature of health data, these initiatives depend on expanding investment into Trusted Research Environments (TRE).
+[@2021c] These are 'data-to-modeler' (DTM) designs where data flows from source (primary, secondary, social care and elsewhere) to a single secure landing zone where research teams write the code to link, clean and analyse the data.
+The insights return to the bedside through clinical guidelines and policy.
+The DTM is also the dominant design pattern in ML4H projects but is fundamentally flawed.
+
+> It is flawed because it imposes a separation between the modeller and the end-user.
+ML4H is not concerned with better guidelines or policy but with better operational and clinical decision making.
+This requires the practitioner to work alongside the end-user because excellent offline model performance provides no guarantee of bedside efficacy.
+Algorithms with inferior technical performance may even provide greater bedside utility.[@the2021; @shah2019] An inverted 'modeler-to-data' (MTD) paradigm was initially proposed to reduce privacy concerns (data are no longer copied and shared[@guinney2018]), but we see important additional value in that it forces 'real-world development' (RW-Dev) and enables the end-user to work with the modeler in rapid-cycle build-test-learn loops.
+
+> ML4H interventions are virtual not physical, and must act by altering the behaviour of clinicians by providing insights for better decisions.
+As such, perfect information provided to the wrong person, or at the wrong moment, cannot be impactful.
+This first pillar of the CDE is the equivalent of an *internal* TRE *within* the healthcare institution.[@guinney2018]
+
+Reference: Guinney J, Saez-Rodriguez J. Alternative models for sharing confidential biomedical data. Nat Biotechnol. 2018 May;36(5):391–2.
+
 2.
@@
 
+Thank you. These references were enormously helpful, and we have included them in the text for Section 2.
+
+> Others have made similar proposals and labelled this as "algorithmic stewardship" or "AI-QI".[@eaneff2020;@feng2022]
+
 3.
@@
 
+- [ ] TODO:https://github.com/docsteveharris/2022-adversarial-penguin/issues/2
+
 4.
@@
 The idea of deploying ensembles with fail-safes is promising as well, and has also been previously described in Feng 2021.
+
+- [ ] TODO: https://github.com/docsteveharris/2022-adversarial-penguin/issues/3
 
@@
 
+- [ ] TODO: https://github.com/docsteveharris/2022-adversarial-penguin/issues/4
+
 6.
@@
 The following papers have proposed methods for doing so, and should be included in the manuscript: Davis, Greevy, Lasko, Walsh, Matheny 2020 and Feng, Gossmann, Sahiner, and Pirracchio 2022.
+
+- [ ] TODO: https://github.com/docsteveharris/2022-adversarial-penguin/issues/4
+<!-- same issue as above -->
 
@@
 
+Thank you.
+We have added the following sentence that signposts some of the literature in this field.
+The argument is best expressed by the following quote from the Nuffield Trust Seminar (Scobie et al), but is clearly complex and not likely to always prevail (Meyer et al)
+
+> When there is genuine equipoise between two commonly used treatments, not doing a point of care trial to find out which has a better risk/benefit profile seems unethical.
+Pragmatic RCTs could enable clinically important questions that can be answered fast based on studies carried out in routine care settings on a full range of participants.
+- Learning Health Care Systems in the NHS. Nuffield Trust Seminar, January 2019 [10]
+
+We have inserted the following sentence (Paragraph 4, Section 5: Continuous clinical evaluation).
+
+> While this is supported by a growing literature on opt-out consent in Learning Healthcare Systems (LHS)[@halpern2018; @london2018a; @scobie2020], progress require careful attention to a range of concerns.[@meyer2019]
+
 8.
@@
 The authors should provide additional context by comparing their ideas to those previously described in these papers.
+
+Thank you. I have pointed the reader to this literature at the beginning of Section: Drug Discovery Parallels.
+
+> We have described a template for a Clinical Deployment Environment that supports the translation of ML4H algorithms from bench to bedside.
+Although the requirements differ, the objective is similar to that for drug development.
+*A similar approach to phasing has previously been proposed for (biomarker) prediction models.*[@pepe2001]
 
@@
 
+This section was is not meant to be correlated in a 1:1 manner with the five pillars.
+We hoped to use this section to convince the reader that algorithm deployment would require the similar effort and infrastructure to that for drug discovery, and that the 'get rich quick' approach which fills the popular and even the scientific press is shortsighted[@bunz2022a], and more thoughtful and careful work is often overlooked.[@sendak2020]
+
+We have added a final paragraph that we hope makes this clear, and similar word of caution in the conclusion.
+
+Section: Drug Discovery Parallels
+
+> The phases of drug development are not meant to be matched 1:1 to the pillars of the CDE described here, but the parallel is drawn to hightlight the effort necessary to see ML4H have an impact on the clinical and operatational decision making in the workplace.
+
+Conclusion Paragraph 4
+
+> They fundamentally are an argument for a professionalisation of ML4H, and a caution against the 'get-rich quick' headlines in the popular and scientific press.[@bunz2022a]
+
 10.
@@
 
-Minor:  
+Thank you.
+We have added a substantial paragraph (below) to the caption and leverage the small numbers (1-5) to explain the process.
+We'd appreciate your and other reviewers advice whether this should remain in the main text or move to the ESM where we have also further expaned the description of data flows in EMAP.
+
+Additional paragraph extending the Figure 1 caption:
+
+> Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.
+EMAP copies each message passing through the integration engine to a PostgreSQL database, the *Immutable Data Store (IDS)* (see Figure 1 and the \[numbers\] in the following sentences that refer to objects in the figure).
+A *message reader*\[2\] processes each live message to an interchange format so that downstream processing is insulated from local HL7 implementation.
+Separately, the *table reader*\[5\] processes historical data (e.g.
+from the reporting database) to the same interchange format.
+Live messages take priority over historical messages in a queue that feeds the *event processor*\[3\].
+This links each message to a patient and a hospital visit, makes appropriate updates for out of order messages, and merges when separate identifiers are recognised to represent the same patient.
+A full audit trail is maintained.
+Each event updates a second live PostgreSQL database, the *User Data Store (UDS)*\[4\].
+> The hospital hosts Jupyter and RStudio servers, and a Linux development environment is provided that allows docker deployment, installation of analysis libraries and frameworks, exposes SSH and HTTPS services, and allows user verification against the hospital active directory.\[5\]\
+> A typical workflow might include investigation and experimentation in a Jupyter Notebook with data from the UDS, then using a small network of docker containers to run the development script, log outputs to a testing database, and report to users via email or a locally hosted web application or dashboard.
+> A fuller explanation is available in the Electronic Supplementary Material [Section 2: EMAP data flows]().
+
+## Minor:
+
 1. Please remove the use of undefined acronyms in the figures (e.g. HL7 and HDR-UK) or provide definitions/citations.
-Check List  
-a. Is the quality of the figures and tables satisfactory?  
-Yes  
-b. Does the reference list cover the relevant literature adequately and in an unbiased manner?  
-Yes  
-c. Does this manuscript refer only to published data? (unpublished or original data is not allowed for this article type)  
-No  
-d. Does the review include a balanced, comprehensive, and critical view of the research area?  
-Yes  
-QUALITY ASSESSMENT:  
-Rigor  
-3  
-Quality of the writing  
-5  
-Overall quality of the content  
-5  
-Interest to a general audience  
-5
+
+Thanks for this careful proof read. The figures are now updated and these acronyms have been removed.

### 2022-06-27 rebuttals/response-reviewer-2.md
first draft of reviewer 2 response

*************************  
Independent Review Report, Reviewer 2  
EVALUATION  
Please list your revision requests for the authors and provide your detailed comments, including highlighting limitations and strengths of the review. If you have additional comments based on Q2 and Q3 you can add them as well.  
Thank you for the opportunity to review this manuscript from Harris, Bonnici, and colleagues regarding the concept of clinical deployment environments. The authors propose this concept as having potential to align development, deployment, and evaluation of electronic health record-based machine learning tools in the healthcare setting.

## Major Strengths
1. The pillars proposed by the authors have face validity and lend important terminology around important considerations within a relatively underdescribed area.  
2. The figures are very nicely done and quite informative.

## Major Limitations
1. The analogy to drug discovery is not a great fit, in my opinion, and its shortcomings reveal an opportunity to better-describe the relationships between ML4H tools and the problems they seek to solve.

- [ ] decide how to play this; see for other examples of using the same analogy "algorithmic stewardship"[@eaneff2020] esp b/c also talks about the team who should advice again making the argument for pillar 1; they also use the drug parallel (trials/hospital formulary/evaluations/stewardship)
> clinicians who are familiar with the language of data, patients, bioethicists, scientists, and safety and regulatory organizations.

2. Some areas of the prose could benefit from increased specificity and/or clarity.

- [ ] idea was to keep the main text narrative and then allow the interested reader to hit the ESM for details with the figures providing the bridge; possible compromise is to expand the captions for each so the flow of the text remains roughly the same??

3. Much of the supplemental content would be better served in the main text.

- [ ] ditto; decide how to play this

## Major Comments
1. The analogy to drug discovery is a poor fit for this essay. One key premise underlying drug discovery activities is that potentially-actionable targets have already been identified through prior research. For example, the decision to seek JAK2-interacting molecules emerged from the understanding that JAK2 regulates important cytokines in a number of disease states. The algorithmic equivalent of this kind of prior knowledge is not apparent to me. Without a clear understanding of (a) a well-articulated problem to be solved and (b) a reasonable mechanism/logic model in which an ML tool – provided it performs well – might affect stakeholder behavior, translational failure seems essentially guaranteed. Said differently, the bake-off between different ML approaches can really only be held once the model (“ligand”)’s “receptor” has been mapped/described, which I would argue happens infrequently at present. As I read the authors’ proposed pillars, I found myself reflecting on the need for this kind of formative work, informed by stakeholders, to ensure that the CDE can be used for properly-specified development, operations, and evaluation.

I might suggest omitting or significantly abbreviating this section, which could open space to insert valuable information from the supplement. I also think a paragraph commenting on the relationship of the CDE to the organization’s “problem identification and specification” apparatus would be a very helpful “bookend” to the implementation and continuous evaluation portions of the essay.

- [ ] ask Tim to think how he wants to play this?
https://github.com/docsteveharris/2022-adversarial-penguin/issues/5

2. Some areas of the prose could benefit from increased specificity and/or clarity. One of the challenges in the healthcare data science space is striking the right balance of technical language to meet a readership with different levels of familiarity to clinical medicine, coding/development, and operations. Some areas of the prose – e.g., lines 63-67 – skew a bit heavy into the data science jargon and might benefit from additional explanations or a bit less jargon. It may be worth the authors having a colleague with less technical familiarity provide some informal copy-editing to help with readability for an audience of varying levels of expertise.

3. The supplement – especially items 1 and 2 – are very nice and highly practical. Some elements of these items could serve to address issue 2 above. Further, the ML-Ops design principles are particularly helpful explicating how code-to-data can address the limitations of data-to-code. I think it would be great if these could be a bit more prominent, ideally in the main manuscript.

Minor Comments   
1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with “This” or “That,” which do not specify what “This” or “That” actually are.  
2. Supplement page 2, typo in the ML-0PS header (zero instead of O)

Thanks for this careful proofread. I double checked. I think this is a quirk of the font that Overleaf is using for the online submission and I can confirm the source document uses `O` not `0`.

i.e.

```markdown
# ML-Ops (for health)
```

not

```markdown
# ML-0ps (for health)
```

### 2022-06-27 rebuttals/response-reviewer-1.md
formatting of response

# Independent Review Report, Reviewer 1

The article proposes a design pattern, Clinical Deployment Environments (CDE) with 05 principles (pillars). A design pattern is a general repeatable and implementable solution which is well demonstrated through the datasheet attached. My overall impression apart from a few specific comments is, that for each of the five principles (pillars), the problem description that they address may be clearly stated in the introduction section. The implementation example in the datasheet describes the first 02 pillars very clearly, however, an implementation example for the rest of the 03 pillars is missing. They should be included to prove the implementation. I would think that the Implementation science is a result of the first two pillars (i.e. RW-Dev and ML-Ops) and not essentially one of the pillars here. The explanation of 'Implementation Science' in the article also does not provide clear requirements for this pillar.

Other specific comments -
- The language style preferred for Frontiers in Digital Health Articles is US English while the article uses the GB English style. A revision is needed to match the editorial requirements
- A review of the state-of-the-art in the area may be included to highlight the gaps
- A thorough check for sentences, the flow of the information, and grammar will help increase the readability of the article. A few of the comments and errors are highlighted in the attached file.
- The diagram in Figure 1. does not match with the attached Image 1
- Suggest moving the figures as part of the datasheet where EMAP and FlowEHR are explained instead of keeping them in the main article
- An overall depiction of CDE and its pillars may be useful
- Datasheet, Page 2, Section 1.4.2, "... aggregation. (201, 2018) Flat (bulk) FHIR is not mature, nor designed for live interactions.(202, 2021)..." : Please update the references. They are not numbered in the bibliography and full information is missing.

### 2022-06-27 rebuttals/response-reviewer-2.md
formatting of response

@@
-*************************  
-Independent Review Report, Reviewer 2  
-EVALUATION  
-Please list your revision requests for the authors and provide your detailed comments, including highlighting limitations and strengths of the review.
-If you have additional comments based on Q2 and Q3 you can add them as well.
+# Independent Review Report, Reviewer 2
+
 Thank you for the opportunity to review this manuscript from Harris, Bonnici, and colleagues regarding the concept of clinical deployment environments.
@@
 The algorithmic equivalent of this kind of prior knowledge is not apparent to me.
-Without a clear understanding of (a) a well-articulated problem to be solved and (b) a reasonable mechanism/logic model in which an ML tool – provided it performs well – might affect stakeholder behavior, translational failure seems essentially guaranteed.
-Said differently, the bake-off between different ML approaches can really only be held once the model (“ligand”)’s “receptor” has been mapped/described, which I would argue happens infrequently at present.
-As I read the authors’ proposed pillars, I found myself reflecting on the need for this kind of formative work, informed by stakeholders, to ensure that the CDE can be used for properly-specified development, operations, and evaluation.
+Without a clear understanding of (a) a well-articulated problem to be solved and (b) a reasonable mechanism/logic model in which an ML tool - provided it performs well - might affect stakeholder behavior, translational failure seems essentially guaranteed.
+Said differently, the bake-off between different ML approaches can really only be held once the model ("ligand")'s "receptor" has been mapped/described, which I would argue happens infrequently at present.
+As I read the authors' proposed pillars, I found myself reflecting on the need for this kind of formative work, informed by stakeholders, to ensure that the CDE can be used for properly-specified development, operations, and evaluation.
 
 I might suggest omitting or significantly abbreviating this section, which could open space to insert valuable information from the supplement.
-I also think a paragraph commenting on the relationship of the CDE to the organization’s “problem identification and specification” apparatus would be a very helpful “bookend” to the implementation and continuous evaluation portions of the essay.
+I also think a paragraph commenting on the relationship of the CDE to the organization's "problem identification and specification" apparatus would be a very helpful "bookend" to the implementation and continuous evaluation portions of the essay.
 
@@
 One of the challenges in the healthcare data science space is striking the right balance of technical language to meet a readership with different levels of familiarity to clinical medicine, coding/development, and operations.
-Some areas of the prose – e.g., lines 63-67 – skew a bit heavy into the data science jargon and might benefit from additional explanations or a bit less jargon.
+Some areas of the prose - e.g., lines 63-67 - skew a bit heavy into the data science jargon and might benefit from additional explanations or a bit less jargon.
 It may be worth the authors having a colleague with less technical familiarity provide some informal copy-editing to help with readability for an audience of varying levels of expertise.
@@
 3.
-The supplement – especially items 1 and 2 – are very nice and highly practical.
+The supplement - especially items 1 and 2 - are very nice and highly practical.
 Some elements of these items could serve to address issue 2 above.
@@
 Minor Comments
-1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with “This” or “That,” which do not specify what “This” or “That” actually are.  
+1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with "This" or "That," which do not specify what "This" or "That" actually are.  
 2. Supplement page 2, typo in the ML-0PS header (zero instead of O)

### 2022-06-27 rebuttals/response-reviewer-3.md
formatting of response

@@
+# Independent Review Report, Reviewer 3
+
 > Summary: I congratulate the authors on writing a wonderful overview on how hospitals can reach the digital maturity to deploy machine learning algorithms to support clinical decision making.
@@
 1.
-I am confused what the terminology “code-to-data” means exactly.
-The authors never really provide a definition of what it means to reverse the typical setup of “data-to-code”.
+I am confused what the terminology "code-to-data" means exactly.
+The authors never really provide a definition of what it means to reverse the typical setup of "data-to-code".
 My first impression was that it meant the ML team would write code that would generate data.
@@
 
-Reference: Guinney J, Saez-Rodriguez J. Alternative models for sharing confidential biomedical data. Nat Biotechnol. 2018 May;36(5):391–2.
+Reference: Guinney J, Saez-Rodriguez J. Alternative models for sharing confidential biomedical data. Nat Biotechnol. 2018 May;36(5):391-2.
 
@@
 For instance, Feng 2022 describes interesting parallels between hospital Quality Improvement units and ML-Ops team (they called this an AI-QI team).
-Eaneff 2020 refers to such teams as “algorithm stewards” to draw links to the concept of “hospital stewards” that is well-understood in the clinical space.
+Eaneff 2020 refers to such teams as "algorithm stewards" to draw links to the concept of "hospital stewards" that is well-understood in the clinical space.
 I think the case for Pillar 2 would be stronger if the authors mention these connections.
@@
 It would be useful for the authors to provide additional explanations on the ethical considerations and why randomization without consent is justified.
-The authors should also highlight the complexity of estimating the ML model’s impact in encouragement designs due to selection bias and noncompliance: see e.g.
+The authors should also highlight the complexity of estimating the ML model's impact in encouragement designs due to selection bias and noncompliance: see e.g.
 Barnard 2011 and Frangakis et al 2002.
@@
 Please expand the caption to explain the major components in the figure.
-For instance, what is the “immutable data store” versus the “user data store”?
-What is the “message queue” vs the “message reader”?
+For instance, what is the "immutable data store" versus the "user data store"?
+What is the "message queue" vs the "message reader"?
 Also, what value add does the reporting DB offer over the live EHR?

### 2022-06-29 rebuttals/response-reviewer-1.md
language/flow

@@
 - The language style preferred for Frontiers in Digital Health Articles is US English while the article uses the GB English style. A revision is needed to match the editorial requirements
+
+Thank you. I believe we have made the necessary changes.
+
 - A review of the state-of-the-art in the area may be included to highlight the gaps

### 2022-06-29 rebuttals/response-reviewer-2.md
language/flow

@@
 
+Thank you. This was helpful. I have asked two colleagues to separately proof the manuscript. They have made a series of small stylistic changes that I have incorporated into the manuscript.
+
+e.g. lines 63-67
+
+Before:
+
+> Given the sensitive nature of health data, these initiatives depend on expanding investment into Trusted Research Environments (TRE).
+[@2021c] Data flows from source (primary, secondary, social care and elsewhere) to a single secure landing zone where research teams write the code to link, clean and analyse the data.
+The insights return to the bedside through clinical guidelines and policy.
+This offline '_data-to-code_' approach is also the dominant design pattern in ML4H projects but is fundamentally flawed.
+
+After
+
+> Given the sensitive nature of health data, these programmes depend on Data Safe Havens (also known as Trusted Research Environments [TRE]).
+[@2021c] TREs are an example of 'data-to-modeler' (DTM) designs where source data (from primary, secondary, and social care) **moves to** a separate, secure landing zone.
+Here research teams link, clean and analyse the data.
+Their insights eventually return to the bedside through clinical guidelines and policy.
+In ML4H, DTM is also the dominant design pattern but is fundamentally flawed.
+
+A detailed diff of these changes is available [here](https://github.com/docsteveharris/2022-adversarial-penguin/pull/7)
+
 3.

### 2022-06-29 rebuttals/response-reviewer-1.md
nearly there

@@
 
-Thank you. I believe we have made the necessary changes.
+Thank you. We have converted converted the article to Word, and applies the US American spell check and grammar, and then converted back to Latex.
 
 - A review of the state-of-the-art in the area may be included to highlight the gaps
+
+Thank you. We have added several new references.
+
+General
+
+1.
+Feng J, Phillips RV, Malenica I, Bishara A, Hubbard AE, Celi LA, et al.
+Clinical artificial intelligence quality improvement: towards continual monitoring and updating of AI algorithms in healthcare.
+npj Digit Med.
+2022 Dec;5(1):66.
+2. Eaneff S, Obermeyer Z, Butte AJ. The Case for Algorithmic Stewardship for Artificial Intelligence and Machine Learning Technologies. JAMA. 2020 Oct 13;324(14):1397–8. 
+3. Lee CS, Lee AY. Clinical applications of continual learning machine learning. The Lancet Digital Health. 2020 Jun 1;2(6):e279–81.
+
+Specific
+<!-- TODO -->
+
 - A thorough check for sentences, the flow of the information, and grammar will help increase the readability of the article. A few of the comments and errors are highlighted in the attached file.
+
+Thank you. We have corrected as follows
+1. Highlight [page 3]:* and aims to answer for ML4H the same challenge that translational medicine brought to bear for drug discovery *and Note [page 3]:* The sentence looks grammatically incomplete
+
+**Re-written**:
+
+> Just as bio-medicine identified translational roadblocks in the bench-to-bedside journey, ML4H also requires a translational infrastructure supporting real-world development not just real-world data.
+
+2. *Highlight [page 7]:* (T2)*and Note [page 7]:* There is no mention or explanation of T2/T1 in the whole article
+
+**Note**
+
+This was defined in paragraph 1 of Section 4.
+
+> This corresponds to the second (T2) arm of translational medicine: implementation science.[@woolf2008b]
+
+3. *Highlight [page 8]:* does not have to relevant in the next *and Note [page 8]:* does not have to be relevant in the next
+
+ **Thank you. Amended.**
+
+ 4.  *Highlight [page 8]:* Excellent and senior patient*and Note [page 8]:* Do you mean excellent patient and public involvement?
+
+ **Yes. Thank you. Amended.**
+
+ 5. *Highlight [page 8]:* At our own institution, *and Note [page 8]:* Would you like to move this part as part of Data Sheet (as implementation example)?
+
+ We would be keen to keep this brief paragraph in the body of the main text.
+Our original draft had the full explanation.
+This short note was meant to provide enough detail to trigger a reader's interest without distracting from the broader emphasis of the article.
+
+ 6.  *Highlight [page 9]:* such binding*and Note [page 9]:* such as binding...
+
+ **Thank you. Amended.**
+
+7. Highlight [page 10]:* where software is (not as) a medical device *and Note [page 10]:* Do you mean Software as Medical Device?
+
+No.
+We were attempting to point out that where software is within a medical device (e.g.
+a cardiac pacemaker) then we already have a good regulatory framework.
+But because ML4H will be deployed in less controlled and varying environment (the EHR) then we need a new approach.
+
+8. *Strikeout [page 10]:* create
+
+ **Thank you. Amended.**
+
 - The diagram in Figure 1. does not match with the attached Image 1
+
+I'm not sure if I understand. We have uploaded 2 copies of each diagram, and in our version I think they are numbered correctly (I have double checked). Could you please have another look. Thanks.
+
 - Suggest moving the figures as part of the datasheet where EMAP and FlowEHR are explained instead of keeping them in the main article
 - An overall depiction of CDE and its pillars may be useful
+
+We're happy to take editorial guidance on this.
+Our inclination is to include the detailed diagrams in the main text and to leave the textual and more technical descriptions of Pillars 1 and 2 in the ESM.
+We think this keeps the main article readable as an overall 'sketch' but allows the interested reader to obtain further detail.
+To this end (as per Reviewer 3), we have added extended captions rather than extend the body text.
+
 - Datasheet, Page 2, Section 1.4.2, "...
@@
 They are not numbered in the bibliography and full information is missing.
+
+Thank you. These were internet (web) references, and we have corrected the missing details.

### 2022-06-29 rebuttals/response-reviewer-2.md
nearly there

@@
 1. The pillars proposed by the authors have face validity and lend important terminology around important considerations within a relatively underdescribed area.
+
+Thank you.
+
 2. The figures are very nicely done and quite informative.
+
+Thank you again.
 
@@
 I also think a paragraph commenting on the relationship of the CDE to the organization's "problem identification and specification" apparatus would be a very helpful "bookend" to the implementation and continuous evaluation portions of the essay.
-
-- [ ] ask Tim to think how he wants to play this?
-https://github.com/docsteveharris/2022-adversarial-penguin/issues/5
+*We thank the reviewer for this comment but feel that changing the manuscript as they suggest would not be the right thing to do.
+The point of the drug development pipeline analogy is to articulate the complexity of the intellectual work required for innovation.
+No-one expects a drug to be developed from scratch in a 3-5 years.
+However, in many cases people are disappointed when healthcare AI algorithms do not yield practice-changing results in the timeframe of a single project.
+The benefit of the Drug Trial Phases is that people are easily able to understand the risk level, the size of trial and the appropriate endpoints.
+This same thinking needs to be related to algorithms.
+Similar points have been made by Pepe (2005) and Ferrante di Ruffano with respect to biomarker discovery.*
+_To specifically address the JAK-2 example, this work would happen in the pre-clinical phase that we describe as algorithm discovery.
+The AI equivalent of JAK-2 (a targetable molecule) is a feature (a representation of data that has some utility in classification or prediction that also has a causal relationship with the phenomenon under study) _
 
@@
 1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with "This" or "That," which do not specify what "This" or "That" actually are.
+
+Thank you. We hope we have corrected this with the proof reading and recommendations (esp. reviewer 1).
+
 2. Supplement page 2, typo in the ML-0PS header (zero instead of O)

### 2022-06-29 rebuttals/response-reviewer-3.md
nearly there

@@
 
-- [ ] TODO:https://github.com/docsteveharris/2022-adversarial-penguin/issues/2
+Thank you. We have revised the paragraph and added the recommended references as follows:
+
+> We argue that model explainability methods [@gunning2019],[@mueller2019],[@vilone2020],[@Linardatos2020] need to be prioritised to help systematise and coordinate the processes of model troubleshooting by developers, risk-management by AI-backed service provider and system-checks by auditors.
+Most AI models that operate as ‘black-box models’ are unsuitable for mission-critical domains, such as in healthcare, because they pose risk scenarios where problems that occur can remain masked and therefore undetectable and unfixable.
+As highlighted in [@Ghassemi 2021],[@Satyapriya2022] explainability methods cannot yet be relied on to provide a determinate answer as to whether an AI-recommendation is correct.
+However, explainability methods that highlight decision-relevant parts of AI representations and for measuring and benchmarking interpretability [@Doshi-Velez2017],[@Hoffman2018] are particularly promising for risk management as they can be used to structure a systematic interrogation of the trade-off between interpretability, model accuracy and the risk of model misbehaviour.
 
@@
 
-- [ ] TODO: https://github.com/docsteveharris/2022-adversarial-penguin/issues/3
+Thank you. We have revised the paragraph and added the recommended references as follows:
+
+> Fail safes, otherwise referred to as selective prediction [@Chow1970],[@ Bartlett 2008],[@Tortorella2000],[@ElYaniv2010],[@Feng2021] should be designed into support systems to pre-empt and mitigate model misbehaviour.
+The European Commission High-Level Expert Group on AI presented the Ethics Guidelines for Trustworthy Artificial Intelligence in April 2019 with recommendations for AI-support systems that continue to maintain human-agency via a human-in-the-loop oversight.
+Prediction models that map patient data to medically meaningful classes are forced to predict within the predetermined set of classes without the option to flag users when the model is unsure of an answer.
+To address this problem, there is good evidence that methods such as Bayesian deep learning and various uncertainty estimates [@abdar2021] can provide promising ways to detect and refer data samples with high probability of misprediction for human expert review [@Leibig2017; @Filos2019; @Ghoshal2020].
+This may even permit less interpretable models to operate when implemented in conjunction with an effective fail-safe system.
 
@@
 
-- [ ] TODO: https://github.com/docsteveharris/2022-adversarial-penguin/issues/4
+Thank you. We have revised the paragraph and added the recommended references as follows:
+
+> As discussed, models that influence the evolution of its own future input data are at risk of performance deterioration over time due to input data shifts [@Davis2017].
+In such cases, continual learning via calibration drift detection and model recalibration [@Davis2020],[@Feng2022] provides a promising solution but remains a challenging paradigm in AI.
+Recalibration with non-stationary incremental data can lead to catastrophic forgetting when the new data negatively interferes with what the model has already learned [@Parisi2019], or a convergence where the model just predicts its own effect and thus should not be updated [@liley2021].
+On the other hand, models can propose poor decisions because of the inherent biases found within the original dataset.
+In this case, dynamic model recalibration is unlikely to be sufficient and larger model revisions may be required.
+Here Pillar 1 (RW-dev) with suitable audit and monitoring via Pillar 2 (ML-Ops) will be required to overcome what would otherwise be a learning process encumbered by regulatory barriers.[@LEE2020]
 
@@
 
-- [ ] TODO: https://github.com/docsteveharris/2022-adversarial-penguin/issues/4
-<!-- same issue as above -->
+Thank you. We have included these additional references in the revised paragraph (quoted under (5) above).
 

### 2022-07-01 rebuttals/response-reviewer-1.md
reviewer 1 complete

@@
 
+**Response:** Thank you.
+We hadn't previously considered the analogy to formal software design patterns ("a design or implementation structure that achieves a particular purpose") but this exactly captures our purpose.
+The manuscript is a conceptual over-view, and the ESM contains the detail where we have made progress ourselves.
+Pillars 3 and 4 are the requirements where we have made the least progress but removing them from the manuscript because of this would send the wrong message.
+We include them to acknowlege their importance and provide a series of useful references.
+
+With regard to Implementation Science, we see this a distinct requirement that depends on Pillars 1 and 2 but is not captured by them.
+Indeed, it requires different personnel and different methodology.
+In most instances we imagine they will deliver their impact through some of form of passive (choice architecture) or active (interruptive alerts) decision support.
+
+We have revised Section 4 to this end.
+
+> Unlike medications, algorithms can only impact health by influencing the behaviour of clinicians and patients.
+This corresponds to the second (T2) arm of translational medicine: implementation science.[@woolf2008b] A well designed, safe, and responsible AI algorithm may still be ineffective if it does not reach a modifiable target on the clinical pathway.[@the2021] *In most instances, this will be via a modification of the choice architecture (passive)[@halpern2018; @main2010] or via interruptive alerts (active) embedded in the EHR.[@main2010]* Effective implementation requires a multi-disciplinary approach including human-computer interaction, behavioural science, and qualitative analysis.[@sendak2020]
+
 Other specific comments -
+
 - The language style preferred for Frontiers in Digital Health Articles is US English while the article uses the GB English style. A revision is needed to match the editorial requirements
 
-Thank you. We have converted converted the article to Word, and applies the US American spell check and grammar, and then converted back to Latex.
+**Response:** Thank you. We have converted converted the article to Word, and applies the US American spell check and grammar, and then converted back to Latex.
 
@@
 
-Thank you. We have added several new references.
+**Response:** Thank you. We have added several new references.
 
@@
 
-Specific
-<!-- TODO -->
+These are alongside additional references in Section 3 (AI safety), and Section 4 (Implementation).
 
@@
 
-Thank you. We have corrected as follows
+**Response:** Thank you. We have corrected as follows
 1. Highlight [page 3]:* and aims to answer for ML4H the same challenge that translational medicine brought to bear for drug discovery *and Note [page 3]:* The sentence looks grammatically incomplete
@@
 
-This was defined in paragraph 1 of Section 4.
+**Response:** This was defined in paragraph 1 of Section 4.
 
@@
 
-No.
+**Response:** No.
 We were attempting to point out that where software is within a medical device (e.g.
@@
 
-We're happy to take editorial guidance on this.
+**Response:** We're happy to take editorial guidance on this.
 Our inclination is to include the detailed diagrams in the main text and to leave the textual and more technical descriptions of Pillars 1 and 2 in the ESM.
@@
 
-Thank you. These were internet (web) references, and we have corrected the missing details.
+**Response:** Thank you. These were internet (web) references, and we have corrected the missing details.

### 2022-07-01 rebuttals/response-reviewer-2.md
reviewer 2 and 3

@@
 
-- [ ] decide how to play this; see for other examples of using the same analogy "algorithmic stewardship"[@eaneff2020] esp b/c also talks about the team who should advice again making the argument for pillar 1; they also use the drug parallel (trials/hospital formulary/evaluations/stewardship)
-> clinicians who are familiar with the language of data, patients, bioethicists, scientists, and safety and regulatory organizations.
+    **Response:** Please see the response under Major Comments (1) below.
 
@@
 
-- [ ] idea was to keep the main text narrative and then allow the interested reader to hit the ESM for details with the figures providing the bridge; possible compromise is to expand the captions for each so the flow of the text remains roughly the same??
+    **Response:** Please see the response under Major Comments (2) below.
 
@@
 
-- [ ] ditto; decide how to play this
+    **Response:** Please see the response under Major Comments (3) below.
 
@@
 As I read the authors' proposed pillars, I found myself reflecting on the need for this kind of formative work, informed by stakeholders, to ensure that the CDE can be used for properly-specified development, operations, and evaluation.
-
 I might suggest omitting or significantly abbreviating this section, which could open space to insert valuable information from the supplement.
 I also think a paragraph commenting on the relationship of the CDE to the organization's "problem identification and specification" apparatus would be a very helpful "bookend" to the implementation and continuous evaluation portions of the essay.
-*We thank the reviewer for this comment but feel that changing the manuscript as they suggest would not be the right thing to do.
+**Response:**
+We thank the reviewer for this comment but feel that changing the manuscript as they suggest would not be the right thing to do.
 The point of the drug development pipeline analogy is to articulate the complexity of the intellectual work required for innovation.
@@
 This same thinking needs to be related to algorithms.
-Similar points have been made by Pepe (2005) and Ferrante di Ruffano with respect to biomarker discovery.*
-_To specifically address the JAK-2 example, this work would happen in the pre-clinical phase that we describe as algorithm discovery.
-The AI equivalent of JAK-2 (a targetable molecule) is a feature (a representation of data that has some utility in classification or prediction that also has a causal relationship with the phenomenon under study) _
+Similar points have been made by Pepe (2005) and Ferrante di Ruffano with respect to biomarker discovery.
+To specifically address the JAK-2 example, this work would happen in the pre-clinical phase that we describe as algorithm discovery.
+The AI equivalent of JAK-2 (a targetable molecule) is a feature (a representation of data that has some utility in classification or prediction that also has a causal relationship with the phenomenon under study)
 
@@
 It may be worth the authors having a colleague with less technical familiarity provide some informal copy-editing to help with readability for an audience of varying levels of expertise.
-
+**Response:**
 Thank you. This was helpful. I have asked two colleagues to separately proof the manuscript. They have made a series of small stylistic changes that I have incorporated into the manuscript.
-
 e.g. lines 63-67
-
 Before:
-
 > Given the sensitive nature of health data, these initiatives depend on expanding investment into Trusted Research Environments (TRE).
@@
 This offline '_data-to-code_' approach is also the dominant design pattern in ML4H projects but is fundamentally flawed.
-
-After
-
+After:
 > Given the sensitive nature of health data, these programmes depend on Data Safe Havens (also known as Trusted Research Environments [TRE]).
@@
 
+    NB:
     A detailed diff of these changes is available [here](https://github.com/docsteveharris/2022-adversarial-penguin/pull/7)
@@
 I think it would be great if these could be a bit more prominent, ideally in the main manuscript.
+**Response:**
+Thank you.
+Our idea was to keep the main text narrative and then allow the interested reader to use the ESM for details with the figures providing the bridge.
+To this end we have (partly in response to Reviewe 3) expanded the captions so the flow of the text remains roughly the same but there is additional detail at hand.
 
-Minor Comments
+## Minor Comments
 1. The manuscript has a number of instances with unspecified pronouns, e.g., sentences starting with "This" or "That," which do not specify what "This" or "That" actually are.  
-
-Thank you. We hope we have corrected this with the proof reading and recommendations (esp. reviewer 1).
+**Response:** Thank you. We hope we have corrected this with the proof reading and recommendations (esp. reviewer 1).
 
 2. Supplement page 2, typo in the ML-0PS header (zero instead of O)  
-
-Thanks for this careful proofread. I double checked. I think this is a quirk of the font that Overleaf is using for the online submission and I can confirm the source document uses `O` not `0`.
-
+**Response:** Thanks for this careful proofread.
+I double checked.
+I think this is a quirk of the font that Overleaf is using for the online submission and I can confirm the source document uses `O` not `0`.
 i.e.
-
 ```markdown
@@
 ```
-
 not
-
 ```markdown

### 2022-07-01 rebuttals/response-reviewer-3.md
reviewer 2 and 3

@@
 
-> Summary: I congratulate the authors on writing a wonderful overview on how hospitals can reach the digital maturity to deploy machine learning algorithms to support clinical decision making.
+Summary: I congratulate the authors on writing a wonderful overview on how hospitals can reach the digital maturity to deploy machine learning algorithms to support clinical decision making.
 The authors describe 5 major pillars that form a Clinical Deployment Environment.
@@
 
+**Response**:
 Thank you for these comments. This was exactly our aim: to provide an schematic overview of the steps that will be required to reach digital maturity given our own experience.
@@
 Please remove/replace this terminology to avoid reader confusion.
-
+**Response**:
 *We have edited the text to make this clearer.
@@
 We take this further and argue that 'data-to-code' ('data to modeler' in their parlance) is flawed because of the separation it imposes between the modellers and the end users.*
-
 *Concretely under Section 1 (Real World Development) we have replaced paragaphs 2-4 with*
-
 > Given the sensitive nature of health data, these initiatives depend on expanding investment into Trusted Research Environments (TRE).
@@
 The DTM is also the dominant design pattern in ML4H projects but is fundamentally flawed.
-
 > It is flawed because it imposes a separation between the modeller and the end-user.
@@
 Algorithms with inferior technical performance may even provide greater bedside utility.[@the2021; @shah2019] An inverted 'modeler-to-data' (MTD) paradigm was initially proposed to reduce privacy concerns (data are no longer copied and shared[@guinney2018]), but we see important additional value in that it forces 'real-world development' (RW-Dev) and enables the end-user to work with the modeler in rapid-cycle build-test-learn loops.
-
 > ML4H interventions are virtual not physical, and must act by altering the behaviour of clinicians by providing insights for better decisions.
@@
 
-Reference: Guinney J, Saez-Rodriguez J. Alternative models for sharing confidential biomedical data. Nat Biotechnol. 2018 May;36(5):391-2.
+    **Reference**: Guinney J, Saez-Rodriguez J. Alternative models for sharing confidential biomedical data. Nat Biotechnol. 2018 May;36(5):391-2.
 
@@
 I think the case for Pillar 2 would be stronger if the authors mention these connections.
-
+**Response**:
 Thank you. These references were enormously helpful, and we have included them in the text for Section 2.
-
 > Others have made similar proposals and labelled this as "algorithmic stewardship" or "AI-QI".[@eaneff2020;@feng2022]
@@
 I think the recommendation for model explainability should be qualified with this fact that the current explainability methods leave much to be desired.
-
+**Response**:
 Thank you. We have revised the paragraph and added the recommended references as follows:
-
 > We argue that model explainability methods [@gunning2019],[@mueller2019],[@vilone2020],[@Linardatos2020] need to be prioritised to help systematise and coordinate the processes of model troubleshooting by developers, risk-management by AI-backed service provider and system-checks by auditors.
@@
 The idea of deploying ensembles with fail-safes is promising as well, and has also been previously described in Feng 2021.
-
+**Response**:
 Thank you. We have revised the paragraph and added the recommended references as follows:
-
 > Fail safes, otherwise referred to as selective prediction [@Chow1970],[@ Bartlett 2008],[@Tortorella2000],[@ElYaniv2010],[@Feng2021] should be designed into support systems to pre-empt and mitigate model misbehaviour.
@@
 Perhaps a better motivating example for dynamic model calibration is provided in Davis, Lasko, Chen, Siew, and Matheny 2017: They show how a ML model for acute kidney injury decays in model calibration.
-
+**Response**:
 Thank you. We have revised the paragraph and added the recommended references as follows:
-
 > As discussed, models that influence the evolution of its own future input data are at risk of performance deterioration over time due to input data shifts [@Davis2017].
@@
 The following papers have proposed methods for doing so, and should be included in the manuscript: Davis, Greevy, Lasko, Walsh, Matheny 2020 and Feng, Gossmann, Sahiner, and Pirracchio 2022.
-
+**Response**:
 Thank you. We have included these additional references in the revised paragraph (quoted under (5) above).
@@
 Barnard 2011 and Frangakis et al 2002.
-
+**Response**:
 Thank you.
@@
 The argument is best expressed by the following quote from the Nuffield Trust Seminar (Scobie et al), but is clearly complex and not likely to always prevail (Meyer et al)
-
 > When there is genuine equipoise between two commonly used treatments, not doing a point of care trial to find out which has a better risk/benefit profile seems unethical.
 Pragmatic RCTs could enable clinically important questions that can be answered fast based on studies carried out in routine care settings on a full range of participants.
-- Learning Health Care Systems in the NHS. Nuffield Trust Seminar, January 2019 [10]
+(via Learning Health Care Systems in the NHS.
+Nuffield Trust Seminar, January 2019 [10])
 
     We have inserted the following sentence (Paragraph 4, Section 5: Continuous clinical evaluation).
-
 > While this is supported by a growing literature on opt-out consent in Learning Healthcare Systems (LHS)[@halpern2018; @london2018a; @scobie2020], progress require careful attention to a range of concerns.[@meyer2019]
@@
 The authors should provide additional context by comparing their ideas to those previously described in these papers.
-
+**Response**:
 Thank you. I have pointed the reader to this literature at the beginning of Section: Drug Discovery Parallels.
-
 > We have described a template for a Clinical Deployment Environment that supports the translation of ML4H algorithms from bench to bedside.
@@
 Finally, continuous model evaluation (Pillar 5) could also be viewed as a Phase IV trial, which is common in drug regulation for monitoring adverse events after drug approval is granted.
-
+**Response**:
 This section was is not meant to be correlated in a 1:1 manner with the five pillars.
 We hoped to use this section to convince the reader that algorithm deployment would require the similar effort and infrastructure to that for drug discovery, and that the 'get rich quick' approach which fills the popular and even the scientific press is shortsighted[@bunz2022a], and more thoughtful and careful work is often overlooked.[@sendak2020]
-
 We have added a final paragraph that we hope makes this clear, and similar word of caution in the conclusion.
@@
     Section: Drug Discovery Parallels
-
 > The phases of drug development are not meant to be matched 1:1 to the pillars of the CDE described here, but the parallel is drawn to hightlight the effort necessary to see ML4H have an impact on the clinical and operatational decision making in the workplace.
@@
     Conclusion Paragraph 4
-
 > They fundamentally are an argument for a professionalisation of ML4H, and a caution against the 'get-rich quick' headlines in the popular and scientific press.[@bunz2022a]
@@
 Also, what value add does the reporting DB offer over the live EHR?
-
+**Response**:
 Thank you.
@@
 We'd appreciate your and other reviewers advice whether this should remain in the main text or move to the ESM where we have also further expaned the description of data flows in EMAP.
-
 Additional paragraph extending the Figure 1 caption:
-
 > Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.
@@
 ## Minor:
-
 1. Please remove the use of undefined acronyms in the figures (e.g. HL7 and HDR-UK) or provide definitions/citations.
-
+**Response**:
 Thanks for this careful proof read. The figures are now updated and these acronyms have been removed.

### 2022-07-02 main.tex
wip
ready to submit
more refs
wip on captions

@@
 
-\title[Translational Digital Medicine]{The Five Pillars of Translational Digital Medicine}
+\title[Clinical Deployment Environments]{Clinical Deployment Environments: The Five Pillars of Translational Machine Learning for Health}
 
@@
 
-% \markdownInput{01-main.md}
-% \markdownInput{01-intro.md}
-% \markdownInput{02-real-world-development.md}
-% \markdownInput{03-ml-ops.md}
-% \markdownInput{04-responsible-ai.md}
-% \markdownInput{05-implementation.md}
-% \markdownInput{06-continuous-evaluation.md}
-% \markdownInput{08-discussion.md}
+% \markdownInput{00-test.md}
+
+\section*{Introduction}
+\markdownInput{10-intro.md}
+
+\markdownInput{20-real-world-development.md}
+\markdownInput{30-ml-ops.md}
+\markdownInput{40-responsible-ai.md}
+\markdownInput{50-implementation.md}
+\markdownInput{60-continuous-evaluation.md}
+
+\section*{Drug discovery parallels}
+\markdownInput{80-drug-development.md}
+
+\section*{Conclusion}
+\markdownInput{81-conclusion.md}
 
@@
 
-\section*{Figure captions}
+\section*{Figures}
 
@@
 \end{center}
-\caption{ 
-Our real-world development is performed on the  Experimental Medicine Application Platform (EMAP).
+\caption{Our real-world development is performed on the  Experimental Medicine Application Platform (EMAP).
 EMAP is a clinical laboratory within which ML4H researchers can iteratively build, test and gather feedback from the bedside.
 It unifies the data and the tools for off-line and online development of ML4H models.
-In brief, EMAP builds a patient orientated SQL database from HL7 version 2 (HL7v2) messages that are being exchanged between hospital systems.
-HL7v2 messages are ubiquitous in health care, and the \textit{de facto} standard for internal communication.
-}\label{fig:1}
+In brief, EMAP builds a patient orientated SQL database from Health Level 7 version 2 (HL7v2) messages that are being exchanged between hospital systems.
+HL7v2 messages are ubiquitous in health care, and the \textit{de facto} standard for internal communication.\newline Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.
+EMAP copies each message passing through the integration engine to a PostgreSQL database, the \textit{Immutable Data Store (IDS)} (see Figure 1 and the \textbf{(numbers)} in the following sentences that refer to objects in the figure).\newline A \textit{message reader} \textbf{(2)} processes each live message to an interchange format so that downstream processing is insulated from local HL7 implementation.
+Separately, the \textit{table reader} \textbf{(5)} processes historical data (e.g.
+from the reporting database) to the same interchange format.
+Live messages take priority over historical messages in a queue that feeds the \textit{event processor} \textbf{(3)}.
+This links each message to a patient and a hospital visit, makes appropriate updates for out of order messages, and merges when separate identifiers are recognised to represent the same patient.
+A full audit trail is maintained.
+Each event updates a second live PostgreSQL database, the \textit{User Data Store} (UDS) \textbf{(4)}.
+The hospital hosts Jupyter and RStudio servers, and a Linux development environment is provided that allows docker deployment, installation of analysis libraries and frameworks, exposes SSH and HTTPS services, and allows user verification against the hospital active directory.
+\textbf{(5)} A typical workflow might include investigation and experimentation in a Jupyter Notebook with data from the UDS, then using a small network of docker containers to run the development script, log outputs to a testing database, and report to users via email or a locally hosted web application or dashboard.
+A fuller explanation is available in the Electronic Supplementary Material (Section 2: EMAP data flows).}\label{fig:1}
 \end{figure}

### 2022-07-02 main.tex
corrected caption and diagram

@@
 EMAP is a clinical laboratory within which ML4H researchers can iteratively build, test and gather feedback from the bedside.
-It unifies the data and the tools for off-line and online development of ML4H models.
+It unifies the data and the tools for off-line and online development of ML4H models (see Figure 1 and the \textbf{(numbers)} in the following sentences that refer to objects in the figure).\newline
 In brief, EMAP builds a patient orientated SQL database from Health Level 7 version 2 (HL7v2) messages that are being exchanged between hospital systems.
-HL7v2 messages are ubiquitous in health care, and the \textit{de facto} standard for internal communication.\newline Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.
-EMAP copies each message passing through the integration engine to a PostgreSQL database, the \textit{Immutable Data Store (IDS)} (see Figure 1 and the \textbf{(numbers)} in the following sentences that refer to objects in the figure).\newline A \textit{message reader} \textbf{(2)} processes each live message to an interchange format so that downstream processing is insulated from local HL7 implementation.
-Separately, the \textit{table reader} \textbf{(5)} processes historical data (e.g.
+HL7v2 messages are ubiquitous in health care, and the \textit{de facto} standard for internal communication.
+Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.\newline
+EMAP copies each message passing through the integration engine to a PostgreSQL database, the \textit{Immutable Data Store (IDS)} \textbf{(1)}.
+A \textit{message reader} \textbf{(2)} processes each live message to an interchange format so that downstream processing is insulated from local HL7 implementation.
+Separately, the \textit{table reader} \textbf{(6)} processes historical data (e.g.
 from the reporting database) to the same interchange format.

### 2022-07-02 main.tex

@@
 HL7v2 messages are ubiquitous in health care, and the \textit{de facto} standard for internal communication.
-Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.\newline
+Rather than multiple pairwise connections between different hospital electronic systems, an integration engine acts as a single hub that routes HL7 messages, and where necessary translates to ensure compatibility.
 EMAP copies each message passing through the integration engine to a PostgreSQL database, the \textit{Immutable Data Store (IDS)} \textbf{(1)}.

### 2022-07-28 body.tex
prep docx for typesetting

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Bold claims and huge investments suggest Machine Learning (ML) will
transform healthcare.\citep{bunz2022a} High impact publications showcase
precision models that predict sepsis, shock, and acute kidney
injury.\citep{komorowski2018, hyland2020, tomasev2019} Outside
healthcare, tech titans such as AirBnB, Facebook, and Uber create value
from ML despite owning `no property, no content and no
cars'.\citep{mcrae2015} Inspired by this, and very much aware of the
flaws and unwarranted variation in human decision
making\citep{braithwaite2020}, government and industry are now laying
heavy bets on ML for Health (ML4H).\citep{nhsx2022, 2021f}

Widespread adoption of electronic health records (EHR) might be thought
a sufficient prerequisite for this ambition. Yet while EHR adoption is
growing at pace\citep{everson2020}, those ML4H models that have reached
the market rarely use the EHR. They are instead embedded in isolated
digital workflows (typically radiology) or medical
devices.\citep{muehlematter2021} Here the context of deployment is
static and self-contained (imaging), or fully specified (devices), and
translation has proved easier to navigate.

In contrast, the EHR is in constant flux. Both the data and the data
model are updating. New wards open, staffing patterns are adjusted and
from time to time major incidents (even global pandemics) disrupt
everything. There are multiple interacting users, and eventually there
will be multiple interacting algorithms, and organizations will face the
ML equivalent of poly-pharmacy.\citep{morse2020a} Algorithms will
require stewards.\citep{eaneff2020} Whilst the aforementioned high
impact prediction models are developed on real-world data, this is not
the same as real-world development. Data are either anonymized and
analyzed offline, or moved out of the healthcare environment into an
isolated Data Safe Haven (DSH) {[}also known as Trusted Research
Environment (TRE){]}.\citep{burton2015} This separation is the first
fracture leading to the oft-cited AI chasm\citep{keane2018} leaving the
algorithms stranded on the laboratory bench.

A future that sees ML4H generate value from the EHR requires an
alternative design pattern. TREs excel at meeting the needs of
population health scientists but they do not have the full complement of
features required to take an ML4H algorithm from bench-to-bedside. Using
drug development as an an analogy, a TRE is custom made for drug
discovery not translational medicine.\citep{woolf2008b}

In this paper, we describe the functional requirements for a Clinical
Deployment Environment (CDE) for translational ML4H. These requirements
map closely to the classical components of translational medicine, but
differ in that algorithms will require ongoing stewardship even after a
successful deployment. The CDE is an infrastructure that manages
algorithms with the same regard that is given to medicines (pharmacy)
and machines (medical physics). Moreover, the value of ML4H will not
just be from externally developed blockbuster models, but will also
derive from specific and local solutions. Our vision of a CDE therefore
enables both \emph{development} and \emph{deployment}.

Our CDE is supported by five pillars:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Real World Development
\item
  ML-Ops for Health
\item
  Responsible AI in practice
\item
  Implementation science
\item
  Continuous evaluation
\end{enumerate}

We describe these pillars below alongside figures and vignettes
reporting early local experience in our journey building this
infrastructure.

\hypertarget{real-world-development}{%
\section{Real World Development}\label{real-world-development}}

Real-world data (RW-Data) means the use of observational data at scale
augmented by linking across multiple data sources to generate insights
simply not available from isolated controlled clinical
trials.\citep{corrigan-curay2018a} The FDA uses data from tens of
millions of patients in its
\href{https://www.sentinelinitiative.org}{Sentinel programme} to monitor
drug safety, and the \href{https://www.opensafely.org}{OpenSafely}
programme in the UK generated impactful insights into COVID-19 within
the first few months of the global pandemic.\citep{williamson2020}

Given the sensitive nature of health data, these initiatives depend on
expanding investment into TREs. \citep{2021c} TREs are an example of
`data-to-modeler' (DTM) designs where data flows from source (primary,
secondary, social care and elsewhere) to a separate, secure landing
zone. Here research teams write the code to link, clean and analyze the
data. Derived insights eventually return to the bedside through clinical
guidelines and policy. To date, DTM is also the dominant design pattern
in ML4H but this approach is fundamentally flawed.

It is flawed because it imposes a separation between the modeller and
the end-user. ML4H is not concerned with better guidelines or policy but
with better operational and clinical decision making. This requires the
practitioner to work alongside the end-user because excellent offline
model performance provides no guarantee of bedside efficacy. Algorithms
with inferior technical performance may even provide greater bedside
utility.\citep{the2021, shah2019} An inverted `modeler-to-data' (MTD)
paradigm was initially proposed to reduce privacy concerns (data are no
longer copied and shared but analyzed in situ\citep{guinney2018}), but
we see important additional value in that it forces `real-world
development' (RW-Dev) and enables the end-user to work with the modeler
in rapid-cycle build-test-learn loops. This first pillar of the CDE is
the equivalent of an \emph{internal} TRE \emph{within} the healthcare
institution.\citep{guinney2018}

RW-Dev has four functional sub-requirements that distinguish it from a
TRE. (1) Firstly, data updates must match the cadence of clinical
decision making. For most inpatient and acute care pathways, decisions
are in real-time (minutes or hours) at the bedside or in the clinic. (2)
Secondly, development using live data must be sandboxed and so the
clinical system responsible for care delivery is protected (3) Thirdly,
privacy must be managed such that teams are able to develop end-user
applications that inevitably display patient identifiable information
(PII) alongside the model outputs: an anonymous prediction is of little
use to a clinician. (4) Fourthly, attention must be paid to developer
ergonomics. Where development and deployment steps are separated
physically (the TRE paradigm) or functionally (via different languages
and technologies), ownership is often split between two different teams.
One team prepares the raw data and develops the model, and another
prepares the live data and deploys the model. We argue instead that the
same team should be able develop \emph{and} deploy. This should
accelerate iteration, reduce cost and increase quality.\citep{2022b}

We illustrate this idea with a description of our local real-world
development platform in \href{fig:1}{Figure 1}, and provide an extended
description in the Electronic Supplementary Material.

\hypertarget{ml-ops-for-health}{%
\section{ML-Ops (for health)}\label{ml-ops-for-health}}

Hitherto in ML4H, the data and the algorithm have been the `celebrity
couple'. State-of-the-art models trained on RW-Data deliver high profile
publications.\citep{tomasev2019, hyland2020} But only a tiny handful
(fewer than 10 studies in a recent high quality systematic review of
nearly 2000 ML4H publications\citep{ben-israel2020}), were prospectively
implemented. The standard offline `data-to-modeler' (DTM) paradigm
described above incurs a significant but `hidden technical debt' that
includes configuration, data collection and verification, feature
extraction, analysis and process tools, compute and storage resource
management, serving infrastructure, and monitoring.\citep{sculley2015}
In fact, the code for the underlying ML model is estimated to be at most
5\% of the total code with the other 95\% as additional code to make the
system work. `Glue-code', `pipeline jungles', and `dead experimental
codepaths' are some of the anti-patterns that make the transition into
production costly and hazardous.\footnote{One infamous example from the
  financial services sector saw a firm lose \$170,000 per second (more
  than \$400m in 45 minutes) when an outdated piece of code leaked into
  production. The firm in question was fined a further \$12m for
  ``inadequate safeguards'' allowing ``millions of erroneous
  orders''.\citep{2013}}

Agencies such as the \href{https://www.fda.gov}{FDA},
\href{https://www.ema.europa.eu/en}{EMA}, and
\href{https://www.gov.uk/government/organisations/medicines-and-healthcare-products-regulatory-agency}{MHRA}
are working toward safety standards for AI and machine learning, but the
majority of these efforts derive from medical devices regulation.
Treating Software as a Medical Device (SaMD) is appropriate where the
algorithms operate within a constant and predictable environment
(e.g.~code embedded within a cardiac pacemaker). But, as already argued,
ML4H models working with the EHR are likely to find themselves operating
in a significantly more complex landscape. This inconstant environment
where algorithms themselves may only have temporary utility has
parallels to the commercial environment exploited so successfully by the
tech giants.

These companies have cultivated an approach to model deployment called
`ML-Ops'. This combines the practices of `DevOps' (a portmanteau of
Software Development plus IT operations)\citep{2022b} that focuses on
the quality and speed with which software moves from concept to
production, with robust data engineering and machine learning. A typical
ML-Ops system monitors raw input data, checks for distribution drift,
provides a feature store to avoid train/serve skew and facilitate
collaboration between teams, and maintains an auditable and monitored
model repository.\citep{john2021} We present a prototype implementation
interacting with the EHRS in Figure 2 (called FlowEHR).

This constant adjustment of algorithms based on their continuously
measured quality and performance needs a workforce as well as a
technology stack. Just as the safe delivery of medicines to the bedside
is the central activity of a hospital pharmacy team, the safe delivery
of algorithms will require the development of similarly skilled and
specialized practitioners, and we should expect to see clinical ML-Ops
departments in the hospital of the future. Others have made similar
proposals and labeled this as ``algorithmic stewardship'' or
``AI-QI''.\citep{eaneff2020, feng2022} Similarly, the FDA is now
proposing `automatic Algorithmic Change Protocols' (aACP) and proposals
have been advanced to guard against gradual deterioration in prediction
quality (``biocreep'').\citep{2021, feng2021}

\hypertarget{responsible-ai-in-practice}{%
\section{Responsible AI in practice}\label{responsible-ai-in-practice}}

Pillars 1 and 2 should engender well designed and well engineered
algorithms, but they do not protect against the unintentional harm that
AI may induce. Algorithms can only learn from a digital representation
of the world that representation in turn cannot encode moral or ethical
standards. Unfair outcomes, discrimination against sub-populations and
bias are all reported shortcomings.\citep{amodei2016} In a dynamic
setting, risk can also arise in the form of degraded predictive
performance over time. Models that modify clinician's behavior alter
patient profiles by design, but predictive success today inevitably
erodes future performance by rendering obsolete the historical patterns
that drove the performance of the original model.\citep{liley2021}
Responsible AI in practice requires a systems approach that preempts and
safe-guards against these potential risks to patients. We highlight
three promising responses to components of this challenge that need to
become part of the risk management approach for ML4H.

\hypertarget{model-explainability}{%
\subsection{Model explainability}\label{model-explainability}}

We argue that model explainability (Explainable Artificial Intelligence
{[}XAI{]}) methods need to be prioritized to help systematize and
coordinate the processes of model troubleshooting by developers,
risk-management by service providers, and system-checks by
auditors.\citep{gunning2019, mueller2019, vilone2020, linardatos2020}
Most AI models that operate as `black-box models' are unsuitable for
mission-critical domains, such as healthcare, because they pose risk
scenarios where problems that occur can remain masked and therefore
undetectable and unfixable. We acknowledge recent
critiques\citep{ghassemi2021, krishna2022a} of explainability methods
that argue the methods cannot yet be relied on to provide a determinate
answer as to whether an AI-recommendation is correct. However, these
methods do highlight decision-relevant parts of AI representations, and
offer promise in measuring and benchmarking
interpretability\citep{doshi-velez2017, hoffman2018}. They are
particularly promising for risk management as they can be used to
structure a systematic interrogation of the trade-off between
interpretability, model accuracy and the risk of model misbehavior.

\hypertarget{model-fail-safes}{%
\subsection{Model fail-safes}\label{model-fail-safes}}

Prediction models that map patient data to medically meaningful classes
are forced to predict without the option to flag users when the model is
unsure of an answer. To address this problem, there is good evidence
that methods such as Bayesian deep learning and various uncertainty
estimates \citep{abdar2021} can provide promising ways to detect and
refer data samples with high probability of misprediction for human
expert review.\citep{leibig2017, filos2019, ghoshal2020} These fail
safes, or selective prediction approaches should be designed into
support systems to preempt and mitigate model
misbehavior.\citep{chow1970, bartlett2008, tortorella2000, elyaniv2010, feng2021}
Of note, the European Commission High-Level Expert Group on AI presented
guidelines for trustworthy AI in April 2019 with such recommendations:
for systems that continue to maintain human-agency via a
human-in-the-loop oversight. This may even permit less interpretable
models to operate when implemented in conjunction with an effective
fail-safe system.

\hypertarget{dynamic-model-calibration}{%
\subsection{Dynamic model calibration}\label{dynamic-model-calibration}}

As discussed, models that influence the evolution of its own future
input data are at risk of performance deterioration over time due to
input data shifts \citep{davis2017}. In such cases, continual learning
via calibration drift detection and model recalibration
\citep{feng2022, davis2020} provides a promising solution but remains a
challenging paradigm in AI. Recalibration with non-stationary
incremental data can lead to catastrophic forgetting when the new data
negatively interferes with what the model has already learned
\citep{parisi2019}, or a convergence where the model just predicts its
own effect and thus should not be updated \citep{liley2021}. On the
other hand, models can propose poor decisions because of the inherent
biases found within the original dataset. In this case, dynamic model
recalibration is unlikely to be sufficient and larger model revisions
may be required. Here Pillar 1 (RW-dev) with suitable audit and
monitoring via Pillar 2 (ML-Ops) will be required to overcome what would
otherwise be a learning process encumbered by regulatory
barriers.\citep{lee2020}

\hypertarget{implementation-science}{%
\section{Implementation science}\label{implementation-science}}

A well designed, safe, and responsible AI algorithm may still be
ineffective if it does not reach a modifiable target on the clinical
pathway.\citep{the2021} Unlike medications, algorithms can only effect
health by influencing the behavior of clinicians and patients. This
translational obstacle parallels the second arm of translational
medicine (T2): implementation science.\citep{woolf2008b} Behavior
change, in most instances, will be via a modification of the choice
architecture (passive)\citep{halpern2018, main2010} or via interruptive
alerts (active) embedded in the EHR.\citep{main2010} Effective
implementation requires a multi-disciplinary approach including
human-computer interaction, behavioral science, and qualitative
analysis.\citep{sendak2020}

We strongly argue that this task will be more difficult if done offline
and in isolation. Pillar 1 crucially permits not just tuning of the
technical performance of the algorithm but rapid build-test-learn cycles
that directly involve the target user and the clinical pathway in
question. This approach will reduce costs and improve impact, sometimes
leading to trade-offs which might appear surprising to those developing
away from the bedside.\citep{shah2019, morse2020a} This efficiency will
again depend on the problem space: where the algorithmic target depends
on information arising from the EHR rather than an isolated device or
image, and where the pathway involves multiple end-users, then
successful implementation will be near impossible if done sequentially
(development then deployment) rather than
iteratively.\citep{sendak2020, connell2019} Academic health science
centres must become design `laboratories' where rapid prototyping at the
bedside crafts the deployment pathway for \emph{effectiveness} (T2)
rather than just efficacy (T1).\citep{woolf2008b}

Investigations to define how system can influence behavior will need
specialist support and tooling. This might require tools embedded within
the user interface to evaluate and monitor user interaction, and capture
user feedback\citep{yusop2017}, or directed implementation
studies.\citep{sutton2020}

Despite the oft cited risks of alert fatigue with Clinical Decision
Support Systems (CDSS)\citep{phansalkar2013}, there is good evidence
that well designed alerts can be
impactful.\citep{park2022, sayres2019, main2010} Overt behavioural
modifications will need a mechanism to explain their recommendation (as
per XAI) or generate trust (see Pillar 5).\citep{mccoy2022} Trust will
possibly be more important where behavior modification is indirect
through non-interruptive techniques (e.g.~re-ordering preference lists
or otherwise adapting the user interface to make the recommended choice
more accessible).

\hypertarget{continuous-clinical-evaluation}{%
\section{Continuous clinical
evaluation}\label{continuous-clinical-evaluation}}

Our analogy with translational medicine breaks down at the evaluation
stage. For drug discovery, evaluation is via a randomized controlled
trial (RCT). Randomization handles unanticipated bias and ML4H should
hold itself to the same standard but of 350,000 studies registered on
\url{ClinicalTrials.gov} in 2020, just 358 evaluated ML4H, and only 66
were randomized.\citep{zippel2021} As usual for ML4H, those RCTs were
not interacting with EHR data. They were evaluations of algorithms
supporting imaging, cataract screening, colonoscopy, cardiotocographs
and
more.\citep{infant2017, titano2018, wang2019, wu2019, lin2019a, turakhia2019a, long2017}

Where the ML4H intervention delivers a novel biological treatment
strategy, then it is appropriate to reach for the full paraphenalia used
in Clinical Trials of Investigational Medicinal Products
(CTIMPs).\citep{komorowski2018} But in many cases, algorithms will be
used to optimize operational workflows and clinical pathways. These
pathways may be specific and contextual rather than generalizable. Poor
external validity is not a critique: an algorithm that is useful or
important in one institution does not have to be relevant in the next
(the `myth of generalizability').\citep{futoma2020} Moreover, the
algorithm is not the same as the patented and fixed active ingredient in
a medicinal product. This is no single point in time nor single host
environment at which it can be declared enduringly effective. This means
that institutions deploying and relying on these tools need a strategy
for rapid continuous clinical and operational evaluation.

This time the EHR may provide an advantage instead of just additional
complexity. Since ML4H algorithms must be implemented through some form
of direct or indirect CDSS, then the next logical step is to randomize
the deployment of those alerts. This in itself is not novel. Randomized
deterministic alerts from CDSS are part of the standard evaluation
toolkit for quality improvement initiatives in at NYU
Langone\citep{horwitz2019}, and for research
elsewhere.\citep{wilson2021} At NYU Langone, such tooling permitted a
small team to deliver 10 randomized trials within a single
year.\citep{horwitz2019}

The final pillar in our CDE uses the same approach for the probabilistic
insights derived from ML4H. Excellent patient and public involvement,
and ethical guidance, will be required to distinguish those algorithms
that require per patient point-of-care consent from those that can use
opt-out or cluster methods. But we think that latter group is large for
two reasons. Firstly, patients are exposed to varying treatment regimes
by dint of their random interaction with different clinicians based on
geography (the healthcare provider they access) and time (staff holidays
and shift patterns etc.). This routine variation in practice is
summarized as the 60-30-10 problem: 60\% of care follows best practice;
30\% is wasteful or ineffective and 10\% is
harmful.\citep{braithwaite2020} Secondly, because the intervention is
informational, there is ethical precedent for patient level
randomization without consent (e.g.~Acute Kidney Injury
alerts).\citep{wilson2021} This hints at a larger and more routine role
for randomization in evaluation of algorithms. This in turn is supported
by a growing\citep{halpern2018, london2018a, scobie2020} but sometimes
conflicting\citep{meyer2019} literature on opt-out consent in Learning
Healthcare Systems (LHS). As such, progress will require careful
attention to a range of concerns.

At our own institution, we have extended this ethical and safety case
one step further, and we are piloting a study design where the
randomization is non-mandatory: a nudge not an order.\citep{wilson2022}
The clinician is explicitly invited to only comply with the
randomization where they have equipoise themselves. Where they have a
preference, they overrule the alert (see Vignette 1 in the Electronic
Supplemental Material).

Embedded randomized digital evaluation should permit rapid evidence
generation, and build the trust needed to support the implementation
described under Pillar 4.

\hypertarget{drug-discovery-parallels}{%
\section{Drug discovery parallels}\label{drug-discovery-parallels}}

We have described a template for a Clinical Deployment Environment that
supports the translation of ML4H algorithms from bench to bedside.
Although the requirements differ, the objective is similar to that for
drug development. A similar approach to phasing has previously been
proposed for (biomarker) prediction models.\citep{pepe2001}

Most ML4H that derives value from the EHR is in the pre-clinical phase.
In drug development, the objective of this phase is to identify
candidate molecules which might make effective drugs. Evaluation is
conducted in vitro. Metrics used to evaluate candidates, such binding
affinity or other pharmacokinetic properties, describe the properties of
the molecule.\citep{lo2018} For ML, the objective is to identify
candidate algorithms, comprising of input variables and model
structures, which might make the core of an effective CDSS. Evaluation
is conducted offline on de-identified datasets. Metrics used to evaluate
candidates, such Area Under the Receiver Operator Curve (AUROC), the F1
score and calibration, describe the properties of the
algorithm.\citep{collins2015}

Phase 1 drug trials are the first time a drug candidate is tested in
humans. They are conducted in small numbers of healthy volunteers. The
aim of the trial is to determine the feasibility of progressing to
trials in patients by determining drug safety and appropriate dosage.
Drug formulation, the processes by which substances are combined with
the active pharmaceutical ingredient to optimize the acceptability and
effective delivery of the drug, is also considered at this stage. Phase
1 ML4H trials are the first time an algorithm candidate is tested within
the healthcare environment. The aim of the trial is to determine the
feasibility of progressing to trials of efficacy by ensuring the
algorithm implementation is safe, reliable and able to cope with
real-world data quality issues. The development of a mechanism to
deliver of algorithm outputs embedded in the clinical workflow is also
be considered at this stage.

Phase 2 drug trials involve recruitment of small numbers patients with
the disease of interest, typically 50 -- 200. The aim is to determine
drug efficacy at treating the disease. Treating clinicians are involved
in so far as they must agree to prescribe the drug for their patients.
The trials are often too short to determine long term outcomes,
therefore surrogate measures such biomarker status or change in tumour
size are used as endpoints.\citep{vannorman2019} Phase 2 ML4H trials
involve recruitment of small numbers of clinicians making the decision
of interest, typically 5 -- 10. The aim is to determine the efficacy of
the algorithm in improving their decisions. Patients are involved in so
far as they must agree to be on the receiving end of these supported
decisions and identifiable data is required. Endpoints are markers of
successful task completion in all cases. Investigations to determine
ways in which the system could be more successful in influencing user
behavior are carried out at this stage. These include usability
analyses, considerations of how well the ML4H/CDSS is integrated into
the overall system and implementation studies to identify how best to
optimize end-user adoption and engagement.\citep{sutton2020}

Phase 3 drug trials involve the recruitment of large numbers of patients
to determine whether a drug is effective in improving patient outcomes.
The gold standard of trial design is a double-blinded randomized
controlled trial (RCT). Phase 3 ML4H trials will require integration of
data from multiple centers for algorithms acting on specific decisions
but inevitably adapted to their local data environment.

The phases of drug development are not meant to be matched 1:1 to the
pillars of the CDE described here: in fact, our argument for
`real-world' \emph{development} deliberately seeks to merge the steps.
But the parallel is drawn to highlight the effort necessary to see ML4H
have an impact on the clinical and operational decision making in the
workplace. Heretofore this effort has been hugely underestimated.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

Even this analogy stops short of the full task of deployment. With drug
development, the universities and the pharmaceutical industry go on to
take advantage of a supply chain to deliver the drug to the hospital
with the necessary quality control and monitoring. Those prescribing and
administering the drug have spent years in training, and are supported
by pharmacists and medication safety experts. And even after the drug is
administered, observation and long term follow-up continue to identify
side-effects and long term hazards.

That network of expertise and infrastructure is largely in place where
software \emph{is within} (not \emph{as}) a medical device, but is only
just being envisioned where the data driving ML4H comes from the EHR.
This distinction needs to be made else the disillusionment with the
promise of ML4H will continue. The technology does have the potential to
change how we deliver health but the methodology alone is insufficient.
The impressive demonstrations of the power of AI and ML to beat humans
in games, and predict protein structures does not mean that these tools
are ready for wide spread deployment.

But we should not be pessimistic. As per author William Gibson, it is
clear that ``The Future Has Arrived --- It's Just Not Evenly
Distributed.'' Beyond healthcare, machine learning has already
demonstrated that it can reliably create value.\citep{mcrae2015} It is
now our responsibility to take those lessons and adapt them for our
patients.

The Five Pillars outlined here are a sketch of that redistribution. They
are born from our local experience (Pillars 1, 2 and 5) and our wider
observations (Pillars 3 and 4). They fundamentally are an argument for a
professionalization of ML4H, and a caution against the `get-rich quick'
headlines in the popular and scientific press.\citep{bunz2022a} We
envision a future where each algorithm is managed in a digital pharmacy
with the same rigor that we apply to medicines. But unlike drugs, some
of these algorithms will have their entire life-cycle, from development
to deployment, managed by the local healthcare provider. Computer vision
tasks that support diagnostic radiology can be partially developed
offline. Components of sepsis prediction tools will transfer from
institution to institution but will need adapting to local clinical
workflows. But there will be opportunity and value for ML4H to optimize
operational tasks that are temporary or specific to that institution.
This means that some development and much of the deployment will require
a suitably trained workforce, and an infrastructure perhaps supported by
these five pillars.
